---
title: "MFE Programming Workshop"
author: "Brett Dunn, Mahyar Kargar, and Dan Yavorski"
date: "Fall 2018"
output:
  beamer_presentation:
    includes:
      in_header: ../slide_header.tex
    keep_tex: false
subtitle: Week 7
slide_level: 2
---

```{r, include=FALSE}
# This code allows us to change the font size
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
```

# `Rcpp` Interface and RStudio

## Why would you want to use another language?
- R is an interpreted language.

  - Low programmer time.
  - A great tool for data munging, statistics, regressions, etc.
  - However, certain tasks in R can be slow. 

- You can greatly improve performance by rewriting key functions in another language, such as C++.

- A good workflow:

    1. Write your program in R.
    2. If the program is too slow, benchmark your code.
    3. Try to speedup any bottlenecks in R.
    4. Convert any remaining bottlenecks to C++.
    

## Bottlenecks in R

Typical bottlenecks that C++ can address include:

- Loops that can’t be easily vectorized because subsequent iterations depend on previous ones. Particularly nested loops.

- Recursive functions, or problems which involve calling functions millions of times.
    - The overhead of calling a function in C++ is much lower than that in R.

- Problems that require advanced data structures and algorithms that R doesn’t provide.
    - Through the standard template library (STL), C++ has efficient implementations of many important data structures, from ordered maps to double-ended queues.


## R Foreign Languange Interfaces

- R provides several built-in functions to interface to other languages:
    - `.C()`, for C and C++ code,
    - `.Call()`, also for C and C++ code,
    - `.Fortran()` for Fortran code,
    - and many more.

- Packages exist to call Java, Python, Julia, and other languages.

- Resources:
    - [Writing R Extensions](https://cran.r-project.org/doc/manuals/r-release/R-exts.html)
    - [Advanced R](http://adv-r.had.co.nz/C-interface.html) by Hadley Wickham

- These interfaces are time-consuming to learn, but the `Rcpp` package provides a much simpler interface.


## `Rcpp`
- The `Rcpp` package is a fantastic tool written by Dirk Eddelbuettel and Romain François (with key contributions by Doug Bates, John Chambers, and J.J. Allaire).

- `Rcpp` makes it very simple to connect C++ to R.

- `Rcpp` makes it easy to pass R objects like vectors, matrices, and lists to C++ and back to R.
    - However, there is overhead in doing this.
    - If you are concerned about speed, consider using the simplest structure.


## Installation of `Rcpp`

- `Rcpp` is a CRAN package, but it requires an additional tool chain of a compiler, linker and more in order to be able to create binary object code extending R.

- On Linux all required components are likely to be present. For macOS, you will need to install the "command line tools" from the `Xcode` developer package (free) and you may want to install a version of [gfortran](https://gcc.gnu.org/wiki/GFortranBinariesMacOS). 
    - See [`Rcpp` FAQ](http://dirk.eddelbuettel.com/code/rcpp/Rcpp-FAQ.pdf) sections 2.10 and 2.16 if you encounter problems with macOS.

- On Windows, users need to install [Rtools](https://cran.r-project.org/bin/windows/Rtools/).

- After you have the tool chain, you can install the `Rcpp` package:
```{r, eval=FALSE}
install.packages('Rcpp')
```

## First Steps
- Three key functions:
    1. `evalCpp()`
    2. `sourceCpp()`
    3. `cppFunction()`
  
## Basic Usage: `evalCpp()`
- `evalCpp()` evaluates a single C++ expression. Includes and dependencies can be declared.
- This allows us to quickly check C++ constructs.

```{r}
library(Rcpp)
evalCpp("2 + 2") # simple test
```

```{r}
evalCpp("std::numeric_limits<double>::max()")
```
  
## Basic Usage: `cppFunction()`
- `cppFunction()` creates, compiles and links a C++ file, and creates an R function to access it.

```{r}
cppFunction("
  int simpleExample() {
    int x = 10;
    return 2.0*x;
}")
simpleExample() # same identifier as C++ function
```


## Basic Usage: `cppFunction()` with C++11
- C++11 is the ISO C++ standard ratified in 2011, and it provides many new features over C++98 or C++03.
- To use C++11 code in R, we add `plugins=c("cpp11")`.
- May need to run this first: `Sys.setenv("PKG_CXXFLAGS"="-std=c++11")`
- See [First steps in using C++11 with Rcpp](http://gallery.rcpp.org/articles/first-steps-with-C++11/).
```{r, size="small"}
Sys.setenv("PKG_CXXFLAGS"="-std=c++11")
cppFunction("
  int exampleCpp11() {
    auto x = 10;
    return 2.0*x;
}",plugins=c("cpp11"))

exampleCpp11() # same identifier as C++ function
```

## Basic Usage: `sourceCpp()`
- `sourceCpp()` is the actual workhorse behind `evalCpp()` and `cppFunction()`. 
- It is described in more detail in the package vignette [Rcpp Attributes](http://dirk.eddelbuettel.com/code/rcpp/Rcpp-attributes.pdf).
- `sourceCpp()` builds on and extends `cxxfunction()` from package inline, but provides even more ease-of-use, control and helpers-–freeing us from boilerplate scaffolding.
- A key feature are the plugins and dependency options: other packages can provide a plugin to supply require compile-time parameters  (`RcppArmadillo`, `RcppEigen`, `RcppGSL`)

## Basic Usage: RStudio
- In R Studio, File / New File / C++ File, and this file is created:

```{Rcpp, eval=FALSE, size='scriptsize'}
#include <Rcpp.h>
using namespace Rcpp;

// This is a simple example of exporting a C++ function to R. You can
// source this function into an R session using the Rcpp::sourceCpp 
// function (or via the Source button on the editor toolbar)...

// [[Rcpp::export]]
NumericVector timesTwo(NumericVector x) {
  return x * 2;
}

// You can include R code blocks in C++ files processed with sourceCpp
// (useful for testing and development). The R code will be automatically 
// run after the compilation.
//

/*** R
timesTwo(42)
*/
```

## Basic Usage: RStudio (Cont’ed)
In R, we ask `Rcpp` to source the function:
```{r}
sourceCpp("./examples/timesTwo.cpp")
```

Now the function is available to use in R:
```{r}
timesTwo(c(12,24))
```

## Basic Usage: RStudio (Cont’ed)
So what just happened?

- We defined a simple C++ function.
- It operates on a numeric vector argument.
- We asked `Rcpp` to ‘source it’ for us.
- Behind the scenes Rcpp creates a wrapper.
- `Rcpp` then compiles, links, and loads the wrapper.
- The function is available in R under its C++ name.

## `Rcpp` Data Types
`Rcpp` defines several C++ data types that correspond with R data types:

- `IntegerVector` for vectors of type integer.
- `NumericVector` for vectors of type numeric.
- `LogicalVector` for vectors of type logical.
- `CharacterVector` for vectors of type character.
- `List` for `list`s.
- `DataFrame` for `data.frame`s.
- `Function` for functions.
- `Environment` for environments.

## Example of `Rcpp::DataFrame`
```{Rcpp, size='footnotesize', eval=F}
#include <Rcpp.h>
using namespace Rcpp; 

// [[Rcpp::export]]
DataFrame timesTwoDF(NumericVector x) {
  // create a new vector y the same size as x
  NumericVector y(x.size());
  // loop through, double x, assign to y
  for(int i = 0; i < x.size(); i++) {
    y[i] = x[i]*2.0;
  }
  // return a data.frame
  return DataFrame::create(
    Named("x")=x,
    Named("y")=y);
}
```

## Example of `DataFrame` (Cont’ed)
Now, we can source the code in R and use the function:
```{r}
sourceCpp("./examples/timesTwoDF.cpp")
x <- 1:3
df <- timesTwoDF(x)
str(df)
```

## Random Number Generation
`Rcpp` simplifies using R's random number generators in C++.

```{Rcpp, size='footnotesize', eval=F}
#include <Rcpp.h>

using namespace Rcpp;

// [[Rcpp::export]]
NumericVector rcppNormals(int n) {
    return rnorm(n);
}
```

In R:
```{r, size='footnotesize'}
sourceCpp("./examples/rcppNormals.cpp")
set.seed(1234)
rcppNormals(5)
```

# More `Rcpp` Examples

## Example: Vector input, scalar output
- One big difference between R and C++ is that the cost of loops is much lower in C++. 
- Let's compare 
```{r, message=FALSE,warning=FALSE}
sumR <- function(x) {
  total <- 0
  for (i in seq_along(x)) {
    total <- total + x[i]
  }
  total
}
```

## Use `cppFunction()`
- `cppFunction()` allows you to write C++ functions in R.
- When you run this code, `Rcpp` will compile the C++ code and construct an R function that connects to the compiled C++ function.

```{r,message=FALSE,warning=FALSE, size='small'}
library(Rcpp)
cppFunction('double sumC(NumericVector x) {
  int n = x.size();
  double total = 0;
  for(int i = 0; i < n; ++i) {
    total += x[i];
  }
  return total;
}')

```


## C++ vs. R Syntax
- The C++ version is similar, but:
    - To find the length of the vector, we use the `.size()` method, which returns an integer. C++ methods are called with `.` (i.e., a full stop).
    - The `for` statement has a different syntax: `for(init; check; increment)`. This loop is initialised by creating a new variable called `i` with      value 0. Before each iteration we check that `i < n`, and terminate the loop if it's not. After each iteration, we increment the value of `i` by     one, using the special prefix operator `++` which increases the value of `i` by 1.
    - **IN C++, VECTOR INDICES START AT 0!** This is a very common source of bugs when converting R functions to C++.
    - Use `=` for assignment, not `<-`.
    - C++ provides operators that modify in-place: `total += x[i]` is equivalent to `total = total + x[i]`. Similar in-place operators are `-=`, `*=`, and `/=`.

## Bechmarking: In C++ loops are much faster than R
- In this example C++ is much more efficient than R:
    - `sumC()` is competitive with the highly optimized built-in `sum()`,
    - `sumR()` is several orders of magnitude slower.

```{r,message=FALSE,warning=FALSE, size="footnotesize"}
library(microbenchmark)
x <- runif(2000)
microbenchmark(sum(x), sumC(x), sumR(x))
```

## Using `sourceCpp`
- So far, we’ve used inline C++ with `cppFunction()`. 
- For real problems, it’s usually easier to use stand-alone C++ files and then source them into R using `sourceCpp()`. 
<!-- - This lets you take advantage of text editor support for C++ files (e.g., syntax highlighting) as well as making it easier to identify the line numbers in compilation errors. -->
- Your stand-alone C++ file should have extension .cpp, and needs to start with:
```{Rcpp, eval=FALSE}
#include <Rcpp.h>
using namespace Rcpp;
```
- And for each function that you want available within R, you need to prefix it with:
```{Rcpp, eval=FALSE}
// [[Rcpp::export]]
```
- In RStudio File/New File/C++ File does these steps for you.

## Compile the C++ code
- To compile the C++ code, use `sourceCpp("path/to/file.cpp")`. 
- This will create the matching R functions and add them to your current session. 
- Note that these functions can not be saved in a `.Rdata` file and reloaded in a later session; they must be recreated each time you restart R. 
- You can embed R code in special C++ comment blocks.
```{Rcpp, eval=FALSE}
/*** R
# This is R code
*/
```

## Example: `meanC` vs. the built-in `mean()`:
<!-- - Running `sourceCpp()` on the following file implements mean in C++ (`meanC`) and then compares it to the built-in `mean()`: -->
```{Rcpp, eval=FALSE}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
double meanC(NumericVector x) {
  int n = x.size();
  double total = 0;
  for(int i = 0; i < n; ++i) {
    total += x[i];
  }
  return total / n;
}
/*** R
library(microbenchmark)
x <- runif(1e5)
microbenchmark(mean(x),meanC(x))
*/
```

<!-- ## Another `Rcpp` Example -->
<!-- 1. In RStudio, File / New File / C++ File. -->
<!-- 2. Enter code in timesTwoRcpp.cpp -->
<!-- ```{Rcpp, eval=FALSE} -->
<!-- #include <Rcpp.h> -->
<!-- // [[Rcpp::export]] -->
<!-- Rcpp::NumericVector timesTwo(Rcpp::NumericVector x) { -->
<!--   return x * 2; -->
<!-- } -->
<!-- ``` -->
<!-- 3. In R, -->
<!-- ```{r, warning=FALSE} -->
<!-- library(Rcpp) -->
<!-- Rcpp::sourceCpp("./examples/timesTwoRcpp.cpp") -->
<!-- timesTwo(c(32,64)) -->
<!-- ``` -->


## Resources
- [Main Rcpp page](http://www.rcpp.org/): `Rcpp` for Seamless R and C++ Integration.
- [`Rcpp` book](http://www.rcpp.org/book/) by Dirk Eddelbuettel. Can download it for free from [SpringerLink](https://link.springer.com/book/10.1007/978-1-4614-6868-4) on the UCLA network.
- [Advanced R](http://adv-r.had.co.nz/) by Hadley Wickham: [Rcpp](http://adv-r.had.co.nz/Rcpp.html) and [R’s C interface](http://adv-r.had.co.nz/C-interface.html) chapetrs.
- [`Rcpp` Gallery](http://gallery.rcpp.org/): Articles and code examples for the `Rcpp` package.
- [Writing R Extensions](https://cran.r-project.org/doc/manuals/r-release/R-exts.html) from CRAN.
- [Extending R](https://www.crcpress.com/Extending-R/Chambers/p/book/9781498775717) by John M. Chambers.


<!-- ## Lab 4 -->
<!-- \begin{center} -->
<!-- \Large Let's work on Lab 4. -->
<!-- \end{center} -->