---
title: "MFE R Programming Workshop"
author: "Brett Dunn and Mahyar Kargar"
date: "Fall 2017"
output:
  beamer_presentation:
    includes:
      in_header: ../slide_header.tex
    keep_tex: false
subtitle: Week 6
slide_level: 2
---

```{r, include=FALSE, echo=FALSE}
#opts_knit$set(progress = TRUE, verbose = TRUE, size="scriptsize")
options(dplyr.print_max=8, dplyr.width=60)
```

# Introduction

## Questions
\begin{center}
\Large Any questions before we start?
\end{center}

## Overview
- `%>%`
- `tidyr`
- `dplyr`

# `%>%`

## The Pipe Operator `%>%`
- The `magnittr` package provides a pipe operator.
- See `vignette("magrittr")`.
- Basic piping:
    - `x %>% f` is equivalent to `f(x)`
    - `x %>% f(y)` is equivalent to `f(x, y)`
    - `x %>% f %>% g %>% h` is equivalent to `h(g(f(x)))`
- The argument placeholder:
    - `x %>% f(y, .)` is equivalent to `f(y, x)`
    - `x %>% f(y, z = .)` is equivalent to `f(y, z = x)`

## Expose the variables with `%$%`
- The `%$%` allows variable names (e.g. column names) to be used in a function.
```{r, warning=FALSE}
library(magrittr)
iris %>%
  subset(Sepal.Length > mean(Sepal.Length)) %$%
  cor(Sepal.Length, Sepal.Width)
```

## Compound assignment pipe operations with `%<>%`
- There is also a pipe operator which can be used as shorthand notation in situations where the left-hand side is being “overwritten”:
```{r, eval = FALSE, warning=FALSE}
iris$Sepal.Length <- 
  iris$Sepal.Length %>%
  sqrt()
```
Use the %<>% operator to avoid the repetition:
```{r, eval = FALSE, warning=FALSE}
iris$Sepal.Length %<>% sqrt
```
- This operator works exactly like %>%, except the pipeline assigns the result rather than returning it.

# `tidyr`

## Hadley Wickham

- [Hadley Wickham](http://had.co.nz) is practically famous in the R world
- He's developed a very large number of useful packages, e.g. `ggplot2` and `lubridate`.
- Today we will look at `dplyr` and `tidyr`.
- Tidy data is data that’s easy to work with: it’s easy to munge (with `dplyr`), visualise (with `ggplot2` or `ggvis`) 
  and model (with `R`’s hundreds of modelling packages). 
- The two most important properties of tidy data are:
    - Each column is a variable.
    - Each row is an observation.
- Check [R for Data Science](http://r4ds.had.co.nz/) book.

## Sample data
- A common problem is a dataset where some of the column names are not names of variables, but values of a variable. 
- Take `table4a`: the column names `1999` and `2000` represent values of the `year` variable, and each row represents two observations, not one.
- `tidyr` is a member of the core `tidyverse`.
```{r,warning=FALSE, message=FALSE}
library(tidyverse)
table4a
```

```{r, include=FALSE, echo=FALSE, size="scriptsize", eval=TRUE, warning=FALSE, message=FALSE}
library(tidyverse)
set.seed(1234)
stocks <- data.frame(
time = as.Date('2009-01-01') + 0:4,
X = rnorm(5, 0, 1),
Y = rnorm(5, 0, 2),
Z = rnorm(5, 0, 4)
)
head(stocks, n = 4)
```

## Bring columns together with `gather()`
- To tidy a dataset like this, we need to **gather** those columns into a new pair of variables. To describe that operation we need three parameters:
    - The set of columns that represent values, not variables. In this example, those are the columns `1999` and `2000`.
    - The name of the variable whose values form the column names. I call that the `key`, and here it is `year`.
    - The name of the variable whose values are spread over the cells. I call that `value`, and here it’s the number of `cases`.

## Bring columns together with `gather()`
- In the final result, the gathered columns are dropped, and we get new `key` and `value` columns. 
```{r}
table4a %>% 
  gather(`1999`, `2000`, key = "year", value = "cases")
```
    

```{r, include=FALSE, echo=FALSE, size="scriptsize", eval=TRUE, warning=FALSE}
stocksm <- stocks %>% gather(stock, price, -time)
stocksm
```


## Split a column with `spread()`
- Spreading is the opposite of gathering. You use it when an observation is scattered across multiple rows. 
- For example, take `table2`: an observation is a country in a year, but each observation is spread across two rows.

```{r}
table2
```

```{r, include=FALSE, echo=FALSE, size="scriptsize", eval=TRUE, warning=FALSE}
stocksm %>% spread(stock, price)
stocksm %>% spread(time, price)
```

## `spread`ing
- To tidy this up, we first analyse the representation in similar way to `gather()`. This time, however, we only need two parameters:
    - The column that contains variable names, the `key` column. Here, it’s `type`.
    - The column that contains values forms multiple variables, the `value` column. Here it’s `count`.

## `spread`ing
- Once we've figured that out, we can use `spread()`, as shown below
```{r}
spread(table2, key = type, value = count)
```


## `spread()` and `gather()` are complements

```{r, include=TRUE, echo=TRUE, size="scriptsize", eval=TRUE, warning=FALSE}
df <- data.frame(x = c("a", "b"), y = c(3, 4), 
                 z = c(5, 6))
df
df %>% spread(x, y) %>% gather(x, y, a:b, na.rm = TRUE)
```

## There's much more

- As usual, read the [vignette](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html) on the CRAN page.
- Also check [Chapter 12](http://r4ds.had.co.nz/tidy-data.html) of R for Data Science book.

# `dplyr`

## Overview of `dplyr`
- `dplyr` provides a grammar of data manipulation.
    - A simple way to interact with data.
- We learn about:
    - tibble structure `tbl`
    - The pipe operator `%>%`
    - Using `dplyr` with databases
- The [`dplyr` introduction vignette ](https://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html) is a good resource.

## `dplyr` and `data.table`
- See this [post](http://stackoverflow.com/questions/21435339/data-table-vs-dplyr-can-one-do-something-well-the-other-cant-or-does-poorly/27840349#27840349).
- Here are my thoughts:
    - For data less than 1 million rows, it is reported that there is not a significant speed difference between the two.
    - For large data that can fit in memory, use `data.table`.
    - For data than cannot fit in memory, you could use `dplyr` with a database backend.
- `dtplyr` is a package to use `dplyr` with `data.table`.
    - It is slower than just using `data.table`.
    

## Data: nycflights13

- To explore the basic data manipulation verbs of `dplyr`, we'll start
with the built in `nycflights13` data frame
- This dataset contains all flights that departed from New York City in 2013.
```{r, include=TRUE, echo=TRUE,  message=FALSE}
library(dplyr)
library(nycflights13)
head(flights,4)
```

## `tbl`s (Tibbles)
- `dplyr` can work with data frames as is, but if you're dealing with
large data, it's worthwhile to convert them to a `tbl_df`.
- A `tbl` will only display the data that will fit in your console.
-`glimpse()` is another nice way to look at the data 
```{r}
flights <- tbl_df(flights)
flights
```

## Single Table Verbs
- `dplyr` aims to provide a function for each basic verb of data manipulation:
- `select()` (and `rename()`)
    -  returns a subset of the columns
- `filter()` (and `slice()`)
    - returns a subset of the rows
- `arrange()` - reorders rows
    - reorders the rows according to single or multiple variables
- `distinct()`
- `mutate()` (and `transmute()`)
    - builds adds new columns from the data
- `summarise()` - calculates summary statistics
    - which reduces each group to a single row by calculating aggregate measures
- `sample_n()` and `sample_frac()`

## Tidy Data
- `dplyr` works best with *tidy* data. i.e. when variables are in columns and observations are in rows.
-  As shown above, you can use `tidyr` to help you create a tidy dataset. 


## Select Columns by Name with `select()`
- `select()` allows you to rapidly zoom in on a useful subset using operations that usually only work on numeric variable positions:
```{r}
# Select columns by name
select(flights, year, month, day)
```

## Select a Range of Columns with `:`
```{r}
# Select all columns between year and day (inclusive)
select(flights, year:day)
```

## An Example of `-(col1:col2)`
```{r}
# Select all columns except those from year to day (inclusive)
select(flights, -(year:day))
```

## `select` Helper Functions
- dplyr comes with a set of helper functions that can help you select groups of variables inside a `select()` call:
- `starts_with("X")`: every name that starts with "X",
- `ends_with("X")`: every name that ends with "X",
- `contains("X")`: every name that contains "X",
- `matches("X")`: every name that matches "X", where "X" can be a regular expression,
- `num_range("x", 1:5)`: the variables named x01, x02, x03, x04 and x05,
- `one_of(x)`: every name that appears in x, which should be a character vector.


## Add New Columns with `mutate()`
```{r, include=TRUE, echo=TRUE, size="scriptsize"}
flights %>% mutate(
       gain = arr_delay - dep_delay,
       speed = distance / air_time * 60) %>% 
  select(year,month,gain,speed)
```


## If you *only* want to keep the new variables, use `transmute()`
```{r, include=TRUE, echo=TRUE, size="scriptsize"}
transmute(flights,
          gain = arr_delay - dep_delay,
          gain_per_hour = gain / (air_time / 60)
)
```

## Filter rows with `filter()`
- `filter()` allows you to select a subset of rows in a data frame.
- The first argument is the name of the data frame.
- The second and subsequent arguments are the expressions that filter
the data frame
- Select all flights on January 1st with:
```{r, include=TRUE, echo=TRUE, size="scriptsize"}
filter(flights, month == 7, day == 20)
```

## Select rows by position

- To select rows by position, use `slice()`

```{r, include=TRUE, echo=TRUE, size="scriptsize"}
slice(flights, 1:10)
```


## Reorder rows with `arrange()`

- `arrange()` works similarly to `filter()` except that instead of filtering or selecting rows, it *reorders* them.

```{r, include=TRUE, echo=TRUE, size="scriptsize"}
arrange(flights, year, month, day)
```

## Use `desc()` to order a column in descending order

```{r, include=TRUE, echo=TRUE, size="scriptsize"}
arrange(flights, desc(arr_delay))
```


## You can rename variables with `rename()`

```{r, include=TRUE, echo=TRUE, size="scriptsize"}
rename(flights, tail_num = tailnum)
```


## Extract unique rows with `distinct`

- A common use of `select()` is to find the values of a set of variables.
- This is particularly useful in conjunction with the `distinct()` verb

```{r, include=TRUE, echo=TRUE, size="scriptsize"}
distinct(select(flights, tailnum))
distinct(select(flights, origin, dest))
```


## Summarise values with `summarise()`

- The last verb is `summarise()`. It collapses a data frame to a single row.
- You can use any function you like in `summarise()` so long as the function can take a vector of data and return a single number.
```{r, include=TRUE, echo=TRUE, size="scriptsize"}
flights %>% summarise(
          delay = mean(dep_delay, na.rm = TRUE))
```



## `dplyr` aggregate functions
- `dplyr` provides several helpful aggregate functions of its own, in addition to the ones that are already defined in R. These include:
    - `first(x)` - The first element of vector `x`.
    - `last(x)` - The last element of vector `x`.
    - `nth(x, n)` - The nth element of vector `x`.
    - `n()` - The number of rows in the data.frame or group of observations that `summarise()` describes.
    - `n_distinct(x)` - The number of unique values in vector `x`.

## Chaining
- The `dplyr` API is functional --- function calls don't have
side-effects.
- You must always save their results. \alert{UGLY}
- To get around this problem, `dplyr` provides the `%>%` operator
- `x %>% f(y)` turns into `f(x, y)`

## Chaining example
```{r, include=TRUE, echo=TRUE, size="scriptsize", eval=TRUE, warning=FALSE, message=FALSE}
flights %>%
group_by(year, month, day) %>%
select(arr_delay, dep_delay) %>%
summarise(arr = mean(arr_delay, na.rm = TRUE),
dep = mean(dep_delay, na.rm = TRUE)) %>%
filter(arr > 30 | dep > 30)
```

## Commonalities

- The syntax and function of all these verbs are very similar:
  - The first argument is a data frame.
- The subsequent arguments describe what to do with the data frame.
- The result is a new data frame
- Together these properties make it easy to chain together multiple
simple steps to achieve a complex result.


## Grouped operations

- These verbs are useful on their own, but they become really powerful when you apply them to groups of observations
- In `dplyr`, you do this by with the `group_by()` function.
- It breaks down a dataset into specified groups of rows.

## Grouped operations (cont.)

- Grouping affects the verbs as follows:
  - grouped `select()` is the same as ungrouped `select()`, except that grouping variables are always retained.
- grouped `arrange()` orders first by the grouping variables
- `mutate()` and `filter()` are most useful in conjunction with window functions (like `rank()`, or `min(x) =` ). They are described in detail in `vignette("window-functions")`.
- `sample_n()` and `sample_frac()` sample the specified number/fraction of rows in each group.
- `slice()` extracts rows within each group.
- `summarise()` is powerful and easy to understand, as described in more detail below.


## `group_by` Example
- For example, we could use these to find the number of planes and the number of flights that go to each possible destination:
```{r, include=TRUE, echo=TRUE, size="scriptsize"}
flights %>%
  group_by(dest) %>%
  summarise(planes = n_distinct(tailnum),
          flights = n())
```


## Multiple table verbs
- `dplyr` implements the four most useful SQL joins:
    - `inner_join(x, y)`: matching `x` + `y`
    - `left_join(x, y)`: all `x` + matching `y`
    - `semi_join(x, y)`: all `x` with match in `y`
    - `anti_join(x, y)`: all `x` without match in `y`

- And provides methods for:
    - `intersect(x, y)`: all rows in both `x` and `y`
    - `union(x, y)`: rows in either `x` or `y`
    - `setdiff(x, y)`: rows in `x`, but not `y`

## Joins from `dplyr` Map to SQL
- `inner_join(x, y)`
    - `SELECT * FROM x JOIN y ON x.a = y.a`
- `left_join(x, y)`
    - `SELECT * FROM x LEFT JOIN y ON x.a = y.a`
- `right_join(x, y)`
    - `SELECT * FROM x RIGHT JOIN y ON x.a = y.a`
- `full_join(x, y)`
    - `SELECT * FROM x FULL JOIN y ON x.a = y.a`
- `semi_join(x, y)`
    - `SELECT * FROM x WHERE EXISTS (SELECT 1 FROM y WHERE x.a = y.a)`
- `anti_join(x, y)`
    - `SELECT * FROM x WHERE NOT EXISTS (SELECT 1 FROM y WHERE x.a = y.a)`

<!-- ## `dplyr` Works with Databases -->
<!-- - See `vignette("databases")`. -->
<!-- - `dplyr` SQL connections: -->
<!--     - `src_sqlite()`, `sql_mysql()`, `sql_postgres()` -->
<!-- - First, we establish a connection: -->
<!-- ```{r, eval=FALSE} -->
<!-- library(DBI) -->
<!-- con <- src_mysql(dbname = "dbname", # database name -->
<!--                  host = "IP_or_WebAddress",  -->
<!--                  port = 3306, # port is 3306 by default -->
<!--                  user = "username", -->
<!--                  password = "password") -->
<!-- ``` -->

<!-- ## Sample `dplyr` Database Commands -->
<!-- ```{r, eval=FALSE} -->
<!-- # View tables in the Database -->
<!-- src_tbls(con) -->
<!-- # Create table references -->
<!-- flights <- tbl(con, "flights") -->
<!-- planes <- tbl(con, "planes") -->
<!-- # Manipulate tables -->
<!-- flights <- left_join(flights, planes, by = "tailnum") -->
<!-- # Collect results -->
<!-- flights <- collect(flights) -->
<!-- ``` -->

## Lab 3
- \Large Let's redo \href{https://github.com/brettdunn/mfeR2017/blob/master/week5/Lab/Lab3.pdf}{Lab 3} with `dplyr`.

