---
title: "MFE R Programming Workshop"
author: "Mahyar Kargar and Dan Yavorsky"
date: "Fall 2018"
output:
  beamer_presentation:
    includes:
      in_header: ../slide_header.tex
    keep_tex: false
subtitle: Week 1
slide_level: 2
---
# Overview

## Goals
- Learn to program in R.

- What does programming mean?
    - Language syntax.
    - Debugging.
    - Finding solutions.
    - Translating math to code.
  
- This is just the beginning; you'll develop these skills throughout the program.

## R as a language
- R is object oriented.
    - Everything is an object and functions operate differently when passed different types of objects.
- R is functional.
    - Everything that happens in R is a function call.
    - You write fewer loops.
    - You write cleaner code.
- R is extendable.
    - Interfaces to other software are part of R.


## R vs C++
 
- Both are useful, and you will use both in the MFE program.
- R is an interpreted language.
    - Low programmer time.
    - A great tool for data munging, statistics, regressions, etc.
    - However, certain tasks in R can be slow (e.g. loops).
- C++ is very fast, but it takes longer to write programs.
- We can use both together!
- A good workflow:
    1. Write your program in R.
    2. If the program is too slow, benchmark your code.
    3. Try to speedup any bottlenecks in R.
    3. Convert any remaining bottlenecks to C++.
    
## R vs MATLAB, Python, ect
- Each language has its own set of strengths and weakness.
- You are better served by learning R and C++ very well, rather than trying to learn R, C++, MATLAB, Python, Julia, SAS, etc.
- The MFE program is just too short.
    - You also need to learn finance!
- Once you are proficient with R and C++, learning other languages is easy.
- Don't become a master of none!

## MATLAB

- If you want to learn MATLAB after learning R, take a look at [R and MATLAB](http://math.umaine.edu/~hiebeler/comp/matlabR.html) by David Hiebeler.

\centerline{
  \includegraphics[width=\textwidth,height=0.7\textheight,keepaspectratio]{./graphics/RandMATLAB.png}
}


## Structure
- I will talk at the beginning of each class.
- For the remainder of the time you will break into your study groups and work on
  programming tasks.
- Tasks are designed to introduce you to the building blocks that will
  be used for course assignments throughout the MFE program.
- This course is a programming course with emphasis on methods for
  finance.
- The key skills will be translating mathematical algorithms into code
  and developing the ability to find helpful resources.

## Questions
\begin{center}
\Large Any questions before we start?
\end{center}


## R Resources: Books

- Introductory:
    - R for Everyone by Jared P. Lander
    - R Cookbook by Paul Teetor (free at [UCLA LearnIT](https://learnit.ucla.edu/))
    - R for Data Science by Hadley Wickham (free as well)
- Intermediate:
    - The Art of R Programming by Norman Matloff
- Advanced:
    - Software for Data Analysis by John Chambers
    - Extending R by John Chambers
    - Advanced R by Hadley Wickham

## Other Resources
- Book series:
    - [Use R!](http://link.springer.com/bookseries/6991) Springer series
        - FYI: Many Springer textbooks are just $25 through <http://link.springer.com/>.  You need to be on campus or signed into the UCLA VPN. You can download the pdfs for free.
    - O'Reilly R Books (free at [UCLA LearnIT](https://learnit.ucla.edu/))
- Built in documentation!
    - `?funcname`
- [Journal of Statistical Software](https://www.jstatsoft.org/index)
- Data science courses on [Coursera](https://www.coursera.org/specializations/jhu-data-science)
- [Data Camp](https://www.datacamp.com/)
- https://www.r-bloggers.com/
- https://twitter.com/rstudiotips
- Google, Stack Overflow, etc.

## R Environment
- First, you need an R distribution.
    - I recommend Microsoft R Open.
    - <https://mran.microsoft.com/download/>
- Second, you need an integrated development environment (IDE) for R.
    - [R Studio](https://www.rstudio.com/products/rstudio/download/preview/) is a fantastic environment to interact with R.
    - Other options:
        - [R Tools for Visual Studio](https://www.visualstudio.com/en-us/features/rtvs-vs.aspx) if you use Visual Studio.
        - [Emacs Speaks Statistics (ESS)](http://ess.r-project.org/) if you use Emacs.
- I am going to assume that you have a working installation of R
  Studio and that you have a basic understanding of how it works.
- I will show you some Visual Studio.
- My focus is going to be on R programming.

## RStudio
\centerline{
  \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio]{./graphics/RStudio.png}
}

## R Tools for Visual Studio

\centerline{
  \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio]{./graphics/RTVS.png}
}

## Course Materials

- https://github.com/brettdunn/mfeR2017
- The materials for this course were created in RStudio, using [R Markdown](http://rmarkdown.rstudio.com/lesson-1.html).
- To create your own RStudio project:
    - File / New Project / Version Control / Git
    - Enter the URL

\centerline{
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio]{./graphics/git.png}
}

# R Basics

## Command Line Interface
- To run a command in `R`, type it into the console next to the `>` symbol and press the Enter key.
```{r}
2 + 3
```
- Up Arrow + Enter repeats the line of code.
- Esc (Windows/Mac) or Ctrl-C (Linux) interrupts a command.

## RStudio
- To start, create a new R Script file.
    - File/New File/R Script
- You can type your commands in the R Script file and run them on the Console.
    - Easy way to save your work.
    - `Ctrl+Enter` sends the line at the cursor to the console.
    - `Ctrl+Shift+S` runs the entire file.
    - Help/Keyboard Shortcuts lists all the available shortcuts.
        - [RStudio Keyboard Shortcuts](https://support.rstudio.com/hc/en-us/articles/200711853-Keyboard-Shortcuts).
        - Check out the multiple cursors.
- For larger tasks with many files, create an R project.
- Visual Studio is similar.

## General Comments
- Make your code easy to read.
- Check out [Hadley Wickham’s](http://adv-r.had.co.nz/Style.html) and [Google's](https://google.github.io/styleguide/Rguide.xml) R Style Guides .
- Comment your code!
    - `#` indicates a comment in R.
    - `Ctrl+Shift+c` in RStudio comments the line.

\centerline{
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio]{./graphics/styleguide.png}
}

## R Packages
- A package is essentially a library of prewritten code designed to accomplish some task or a collection of tasks.
- R has a huge collection of user-contributed packages.
    - Warning: Not all packages are of the same quality.
    
    
\centerline{
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio]{./graphics/cran.png}
}

## R Packages
- Installing a packages:
    - `Ctrl+7` in RStudio accesses the packages pane
    - You can also type `install.packages("packageName")`
- Uninstalling a package:
    - `remove.packages("packageName")`
- Loading packages:
    - `require(packageName)` or `library(packageName)` loads a package into R
    - The difference is that `require ` returns `TRUE` if the package loads or `FALSE` if it doesn't.
- Unloading packages
    - `detach(package:packageName)`
- If two packages have the same function name use two colons:
    -`package1::func` or `package2::func`

## Getting Help in R

- To get help on a function, use `?`.
- The `example` function runs the examples contained in the help file.
- To run a search through R's documentation, use `??`.
- To get help on a package, type `help(package="packageName")`
```{r, eval=FALSE}
?seq  # pulls up the help page
example(seq)  # runs the examples in R
??"normal distibution"  # runs a search
help(package = "xts")  # gets help on the xts package
?'+'  # gets help on the + function
```


## Variables
- Unlike C++, R does not require variable types to be declared.
- A variable can take on any data type.
- A variable can also hold any R object such as a function, the result of an analysis, a plot, etc.
- Variable assignment is done with  `<-` (`Alt+-` in RStudio).
    - `=` works, but there are reasons to prefer `<-`.
- We can remove variables (e.g. to free up memory) with the `rm` function. 
    - `gc()` runs garbage collection.
    - `rm(list=ls())` clears the workspace.

```{r}
x <- 2  # x is a pointer
x  # the same output as print(x)
rm(x)  # removes x
```

## Data Types
- There are many different data types in R.
- The four main types of data most likely to be used are:
    1. `numeric`
    2. `character` (string)
    3. `Date/POSIXct` (time-based)
    4. `logical (TRUE/FALSE)`

- The data type can be checked with the `class` function
```{r}
x <- as.Date("2010-12-21")
class(x)
```

## Casting
```{r}
x <- "2010-12-21"
class(x)
x
x <- as.Date(x)
class(x)
x
```

## More Casting
```{r}
x <- as.numeric(x)
class(x)
is.numeric(x)
x  # number of days since Jan 1, 1970
```

## Even More Casting
```{r}
x <- as.integer(x)  # x <- 14964L assigns an integer
class(x)
is.integer(x)
is.numeric(x)  # R promotes int to numeric as needed
4L / 5L
```

## Logicals
```{r}
# TRUE == 1 and FALSE == 0
x <- TRUE  # TRUE, FALSE, T, F are logicals
is.logical(x) 
5 == 5  # != tests for inequality
"a" < "b"  # works on characters as well
```


# Vectors


## Vectors
- A `vector` is a collection of elements, all of the *same* type.
- In R, a `vector` does not have a dimension attribute.
    - There is no difference between a row vector and a column vector.

- We will learn about:
    - Recycling
        - The automatic lengthening of vectors.
    - Filtering
        - The extraction of subsets of vectors.
    - Vectorization
        - Where functions are applied element-wise to vectors.
    

## Vectors and Assignment
- Assigning values to variables can be done with `<-`.
- Often, we create vectors using the `c()` function.
    - The "c" stands for combine because the arguments into a vector.
    
```{r}
x <- c(1, 2, 3, 4)
x
y <- c(x, 5, 6)
y
```

## Creating Vectors with `seq` and `rep`
- Both `seq` and `rep` are useful functions for generating vectors.
- See `?seq` and `?rep` for details
- `seq` is also useful in loops
- `1:10` is the same as `seq(1,10,1)`
```{r}
x <- seq(from = 1, to = 10, by = 2) 
x
y <- rep(c(1, 2), times = 3)
y
rep(c(1,2), each=2)
```

## Obtaining the Length of a Vector
- `length()` returns the vector length
```{r}
x <- c(TRUE, FALSE, TRUE, FALSE)
length(x)
x <- c()     # x is NULL
1:length(x)  # that could mess you up in a for loop
seq(x)       # a safe way to loop through a vector
```

## Accessing Elements of Vectors
- Elements can be accessed using `[]`
    - Help on the `[` function can be found by typing `?'['`
- Unlike C/C++, R indexing starts at 1, not 0.
- The `[` function can take a vector as an arguments.
```{r}
x <- c("a", "b", "c", "d")
x[1]                    # access the first element
x[c(1, 3)]              # access elements 1 and 3
x[c(TRUE, FALSE, TRUE, FALSE)]  # second way
```

## NULL and NA
- `NULL` is the non-existent value in R.
- `NA` is the missing place holder.
```{r}
x <- 5:8
x[2] <- NA
x
y <- NULL
length(y)
```

## Names of Vector Elements
- You can give names to elements of vectors, and you can access elements by their name.
- The function `as.vector` removes the names from a vector.
```{r}
x <- 1:3
names(x) <- c("A","B","C")
x <- c(A=1, B=2, C=3 )  # another way
x["B"]
as.vector(x)            # the names are removed
```

## Recycling
- When applying an operation to two vectors that requires them to be the same length, R automatically *recycles* the shorter one, until it is long enough to match the longer one.
- Be careful with and aware of this behavior!
- In some cases it is useful, others confusing.
```{r}
# the shorter vector will be recycled
c(2, 4, 6) + c(1, 1, 1, 2, 2, 2)
# this is the same as
rep(c(2, 4, 6), 2) + c(1, 1, 1, 2, 2, 2)
```

## Logical Operators
- R has several logical operations that act on vectors.
- `!`, `==`, `!=`, `&`, `&&`, `|`, `||`, `xor()`, `any()`, `all()`, `>`, `>=`, `<=`, `<`
```{r}
x <- c(TRUE,FALSE,TRUE)
y <- c(TRUE,FALSE,FALSE)
x == y
!x
```

## Logical Operations (2)
- `&&`, `||`, `any()`, and `all()` return a length-one vector.
- The shorter forms are vectorized, meaning they can return a vector:
```{r}
x <- c(FALSE,TRUE,TRUE)
y <- c(TRUE,TRUE,FALSE)
x & y
```
- The longer form evaluates left to right examining only the first element of each vector:
```{r}
x && y
```

## Filtering
- We select subsets of vectors with vectors of logicals.
```{r}
x <- 1:5
y <- c(TRUE,FALSE,TRUE,FALSE,TRUE)
x[y]
```

## Filtering (2)
- Filtering amounts to generating filtering indices (i.e. vectors of logicals).
```{r}
x <- c(5, 2, -3, 8)
idx <- x*x > 8  
idx
# another way
">"(x*x, 8)
```

## Assigning to a Filter
- You can assign elements to the subsets.
    - This allows you change elements that meet certain criteria.
```{r}
x <- 1:6
x[x <= 2] <- NA
x
```

## Filtering with `subset()`
- The `subset` function filters and removes any `NA`s.
```{r}
x <- c(3, 1:5, NA, 79)
x
x[x > 4]
subset(x, x > 4)
```
## Filtering with Indices
- Select using the row/column number.
- Exclude entries with a negative vector.
```{r}
x <- 10:20
x[c(1:2, 6:7)]
x[-c(1,3)] # removes 1st and 3rd obs
```
## The Selection Function `which()`
- `which()` gives us the *position* in a vector where a condition occurs.
```{r}
x <- c(3, 1:5, NA, 79)
x > 4
which(x > 4)
```
- See https://stackoverflow.com/questions/6918657/whats-the-use-of-which.

## `%in%`
- `%in%` returns a logical vector indicating if there is a match or not for its left operand.
```{r}
1:5 %in% c(1,3)
```


## Vectorization: Functions on Vectors
- R functions typically operate on vectors.
- Often, there is an argument to ignore missing data.
```{r}
x <- c(1:1000, NA)
mean(x)
mean(x, na.rm = TRUE)
log(x)[998:1001]
```


# Lists

## Creating Lists
- A `list` is a structure that combines objects of *different* type and length.
- You can create a `list` where the elements are of type `list`.
```{r}
element1 <- 1:5
element2 <- matrix(1:6, nrow=2)
mylist <- list(el1=element1, el2=element2)
mylist
```

## Accessing Elements of Lists
- We can access a list component in several different ways.
```{r}
mylist <- list(A=1, univ=c("UCLA", "USC"),
               mymat=matrix(1:4, nrow=2))
mylist[[1]]    # first way
mylist[["A"]]  # second way
mylist$A       # third way
```

## Removing Components of Lists
- We can delete a component of a list by setting it to `NULL`.
```{r}
mylist <- list(A=1)
mylist$B <- c(1, 2)  # adds a component to a list
mylist
mylist$A <- NULL
mylist
```

## Subsetting Lists
- Subsets of lists are done with single `[]`.
    - A single `[]` returns a sublist of the original list

```{r}
mylist <- list(A=1, univ=c("UCLA", "USC"),
               mymat=matrix(1:4, nrow=2))
# this returns a list because of the single []
mylist[c(1,3)]
```

## Lists of Objects
- Lists can store all types of objects.
```{r}
l <- list(a = matrix(1:4, nrow=2),
          b = list(A=1:10,B=20:30))
l[["a"]]
l[[2]][[1]]
```

## Simplfying vs Preserving Subsetting
- Note that `[[` simplifies the result, returning a vector (`str` displays the structure of an R object):
```{r}
str(mylist[[2]])
```
- But `[` is preserving subsetting and returns a list:
```{r}
str(mylist[2])
```


## Applying Functions to a List with `lapply`
- `lapply` implicitly loops over each list element and applies a function.
```{r}
mylist <- list(A=1:10,B=2:17,C=745:791)
lapply(mylist,mean)
```

## An Example of `lapply`
- From `?lapply`: `lapply(X, FUN, ...)` returns a list of the same length as X, each element of which is the result of applying FUN to the corresponding element of X.
```{r}
l <- c("A","B","B","A","A","B")
lapply(c("A","B"), function(letter) which(l==letter))
```

## Investments Problem Set 1
- You will submit a list of answers.
```{r}
hw <- source("./lecture1p.R")$value
str(hw)
```



