---
title: "MFE R Programming Workshop"
author: "Brett Dunn and Mahyar Kargar"
date: "Fall 2017"
output:
  beamer_presentation:
    includes:
      in_header: ../slide_header.tex
    keep_tex: false
subtitle: Week 2
slide_level: 2
---
```{r, include=FALSE}
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
```

# Introduction

## Questions
\begin{center}
\Large Any questions before we start?
\end{center}

## Overview of Week 2
- Functions
- Optimization
- Control Statements
- Lab

# Functions

## Everything that happens in R is a function call
```{r}
`<-`(mynumber, 3)
print(mynumber)
`+`(3,4)
a <- `:`(11,20)
`[`(a,5)
```
## Function Definitions
```{r}
myfunc <- function(x) x^2
myfunc(10)
```
- The last value evaluated is what is returned by the function.
- You can also write `return(x^2)`.
    - This is useful if you want to break out of the function early.


## Scope Rules for Functions

- Variables defined inside a function are local to that function.
```{r}
myfunc <- function(x) {
    N <- 10
    N * x^2  # return(N*x^2) is optional
}
myfunc(10)
# You can't access N out here
```

## Default Arguments
- `R` makes frequent use of default arguments.
- Default arguments are initialized with a default value unless the user specifies an override.
```{r}
g <- function(x, y=2, z=2) x*y*z
g(2)
g(2, 1)
g(2, 1, 1)

```
# `%>%`

## The Pipe Operator `%>%`
- Pipes are a powerful tool for clearly expressing a sequence of multiple operations.
- The `magrittr` package provides a pipe operator.
- See `vignette("magrittr")`.
- Basic piping:
    - `x %>% f` is equivalent to `f(x)`
    - `x %>% f(y)` is equivalent to `f(x, y)`
    - `x %>% f %>% g %>% h` is equivalent to `h(g(f(x)))`
- The argument placeholder:
    - `x %>% f(y, .)` is equivalent to `f(y, x)`
    - `x %>% f(y, z = .)` is equivalent to `f(y, z = x)`

## An Example of `%>%`
```{r, warning=FALSE}
library(magrittr)
x <- 1:10
x %>% mean
y <- c(x, NA)
y %>% mean(na.rm=TRUE)
```

## Expose the variables with `%$%`
- The `%$%` allows variable names (e.g. column names) to be used in a function.
```{r, warning=FALSE}
iris %>%
  subset(Sepal.Length > mean(Sepal.Length)) %$%
  cor(Sepal.Length, Sepal.Width)
```

## Compound assignment pipe operations with `%<>%`
- There is also a pipe operator which can be used as shorthand notation in situations where the left-hand side is being “overwritten”:
```{r, eval = FALSE, warning=FALSE}
iris$Sepal.Length <- 
  iris$Sepal.Length %>%
  sqrt()
```
- Use the %<>% operator to avoid the repetition:
```{r, eval = FALSE, warning=FALSE}
iris$Sepal.Length %<>% sqrt
```
- This operator works exactly like %>%, except the pipeline assigns the result rather than returning it.

# Control Statements

## WARNING!
- In general, loops are *slow* in `R`.
- Vectorized code is faster.
- Loops are often a good place to write a R function in C++.
    - We will cover this later.
    
## Looping vs Vectorization
```{r, size = 'footnotesize'}
library(microbenchmark)
sq <- function(x) {
  n <- length(x)
  res <- rep(NA, n)
  for(i in 1:n) res[i] <- x[i]^2
  res
}
x <- 1:1000
microbenchmark(
  sq(x),
  x^2
)
```

## For loops (1)
- A `for` loop iterates over an index, provided as a vector.
- To iterate over the length of a vector `x`, we can either use `1:length(x)` or `seq(x)`.
    - `seq(x)` protects against zero-length vectors.
```{r}
x <- c(1:5)
y <- NULL  # we need to initialize an empty vector
for(i in seq(x)) {  # safer than 1:length(x)
   y[i] <- x[i] + 2
}
y
```


## For loops (2)
- Another nice way to make a `for` loop.
```{r}
x <- c(2:4)
for(i in x) {
   print(i + 2)
}
```


## While loops
- A `while` loop runs the code inside the braces repeatedly as long as the tested condition proves `TRUE`.
```{r}
x <- c(1:5)
y <- NULL
i <- 1
while(i <= length(x)) {
   y[i] <- x[i] + 2
   i <- i + 1
}
y
```


## If Statements
- If statements operate on length-one logical vectors.
- Syntax:
    - `if(cond1=true) { cmd1 } else { cmd2 }`
```{r}
myabs <- function(x) {
  if(x < 0) {
    myabs <- -x
  }
  x
  return(myabs)
}
x <- -10
myabs(x)
```

## Ifelse Statements
- Ifelse statements operate on vectors of variable length.
- Syntax:
    - `ifelse(test, true_value, false_value)`
```{r}
x <- 1:10 # Creates sample data
ifelse(x<5 | x>8, x, 0)
```

# Optimization in R

## Overivew
- R has many tools to solve optimization problems.
- See [CRAN Task View: Optimization and Mathematical Programming](https://cran.r-project.org/web/views/Optimization.html).
- [Continuous Global Optimization in R](https://www.jstatsoft.org/article/view/v060i06/v60i06.pdf)

## One Dimensional Root (Zero) Finding
- The function `uniroot` searches an interval for a root (i.e., zero) of the function `f` with respect to its first argument.
- Suppose we want $x$ such that $5 - e^{x^2} = x^2$.
- We convert this to the root finding problem $5-e^{x^2}-x^2 = 0$.
```{r}
f <- function(x) {
  5 - exp(x^2) - x^2
}
root <- uniroot(f, c(0, 2), tol = 10^-8)
root$root
f(root$root)
```

## General-Purpose Optimization with `optim`
- [Rosenbrock Banana function](https://en.wikipedia.org/wiki/Rosenbrock_function)
```{r}
fr <- function(x) {   ## Rosenbrock Banana function
    x1 <- x[1]
    x2 <- x[2]
    100 * (x2 - x1 * x1)^2 + (1 - x1)^2
}
optim(c(-1.2,1), fr)$par
```

## Investments Problem Set 2
- You will use `R` to calculate yield-to-maturity.


# Lab 1

## Reading in Data for Lab 1
- `read.table` is the basic function to read in tabular data.
- `read.csv` is a special case of `read.table`.
    - As usual see `?read.table`.
    - Often you want to set `stringsAsFactors = FALSE`.
```{r}
optdata <- read.csv(file="./lab/optionsdata.csv",
                header = T, stringsAsFactors = FALSE)
head(optdata, 3)
```
- `optdata` is a `data.frame`, a specialized type of `list`
- `write.csv` writes data to a .csv file.

## Lab 1
\begin{center}
\Large Let's work on Lab 1.
\end{center}