---
title: "MFE R Programming Workshop"
author: "Brett Dunn and Mahyar Kargar"
date: "Fall 2017"
output:
  beamer_presentation:
    includes:
      in_header: ../slide_header.tex
    keep_tex: false
subtitle: Week 4
slide_level: 2
---
# Introduction

## Questions
\begin{center}
\Large Any questions before we start?
\end{center}

## Overview of Week 4
- Strings
- Dates
- Plotting with Dates
- Time-Series Data in `R` (`xts`)
- Plotting in `R`
- Lab

# Strings

## Strings
- A string is a sequence of characters.
- In `R`, a string falls in the `character` class.
```{r}
mystring <- "Hello"
str(mystring)
```
- Character vectors are created like numeric vectors.
```{r}
myvec <- c("Hello", "Goodbye")
str(myvec)
```

## Manipulating Strings
- `R` provides many functions to manipulate strings.
    - `grep()`: Searches for a substring, like the Linux command of the same name.
    - `gsub()`: Replaces all matches of a substring with another string.
    - `nchar()`: Finds the length of a string.
    - `paste()` and `paste0()`: Assembles a string from parts.  
    - `sprintf()`: Returns a character vector containing a formatted combination of text and variable values.
    - `substr()`: Extracts a substring.
    - `strsplit()`: Splits a string into substrings.
- Hadley Wickham's `stringr` package provides additional functions for using regular expressions and examining text data. See more [here](https://github.com/rstudio/cheatsheets/raw/master/strings.pdf).

## `grep()`
- The call `grep(pattern, x)` searches for a specified substring pattern in a vector `x` of strings.
- If `x` has `n` elements—that is, it contains `n` strings—then `grep(pattern, x)` will return a vector of length up to `n`.
- Each element of this vector will be the index in `x` at which a match of pattern as a substring of `x` was found.
```{r}
grep("Pole",c("Equator","North Pole","South Pole"))
```

## `gsub()`
- The call `gsub(pattern, replacement, x)` searches for a specified substring pattern in a vector `x` of strings and replaces it with the provided replacement string.
- If `x` has `n` elements, `gsub()` will return a string vector of length to `n`.
- If the substring is not found, it returns the original string.
```{r}
gsub(" Pole","",c("Equator","North Pole","South Pole"))
```


## `nchar()`
- The call `nchar(x)` finds the length of a string `x`.
```{r}
nchar("South Pole")
```

## `paste()`
- The call `paste(...)` concatenates several strings, returning the result in one
long string.
```{r}
paste("North","and","South","Poles")
paste("North","Pole",sep="")
# paste0 is same as sep="" (more efficient)
paste0("North","Pole") == paste("North","Pole",sep="") 
```

## `sprintf()`
- The call `sprintf(...)` assembles a string from parts in a formatted manner.
- Similar to the `C` function `printf`.
```{r}
i <- 8
sprintf("the square of %d is %d",i,i^2)
```

## `substr()`
- The call `substr(x,start,stop)` returns the substring in the given character position range `start:stop` in the given string `x`.
```{r}
substr("Equator",start = 3,stop = 5)
```

## `strsplit()`
- The call `strsplit(x,split)` splits a string `x` into a list of substrings based on another string `split` in `x`.
```{r}
strsplit("10-05-2017",split="-")
```

## Example: Creating File Names
- Suppose we want to create five files, `q1.pdf` through `q5.pdf`, consisting of
histograms of 100 $N\left(0,i^2\right)$ random variables. We could execute the following
code:
```{r, eval=FALSE}
for (i in 1:5) {
  fname <- paste("q",i,".pdf")
  pdf(fname)
  hist(rnorm(100,sd=i))
  dev.off()
}
```

# Dates


## Why do we need date/time classes?

\centerline{
  \includegraphics[width=\textwidth,height=0.85\textheight,keepaspectratio]{./railroad_time.jpg}
}


## Date Classes in `R`
- `Date` is in `yyyy-mm-dd` format and represents the number of days since Jamuary 1, 1970
- `POSIXct` represents the (signed) number of seconds since Jamuary 1, 1970 (in the UTC time zone) as a numeric vector.
- `POSIXlt` is a named list of vectors representing `sec`, `min`, `hour`, `mday`, `mon`, `year`, time zone par maters, and a few other items.
```{r}
x <- Sys.time()  # clock time as a POSIXct object
x; as.numeric(x)
```

## Creating Dates
- Typically, dates come into `R` as character strings.
- By default, `R` assumes the string is in the format `yyyy-mm-dd` or `yyyy-mm-dd`
```{r}
mychar <- "2017-10-05"
mydate <- as.Date(mychar)
str(mydate)
```

## Date Formats
- `R` can parse many other types of date formats.
- See `?strptime` for details.
```{r}
mychar <- "October 5th, 2017"
mydate <- as.Date(mychar, format = "%B %eth, %Y")
str(mydate)
```

## Extract Parts of a Date Object
```{r}
mydate <- as.Date("2017-10-05")
weekdays(mydate)
months(mydate)
quarters(mydate)
```

## Generate Regular Sequences of Dates
```{r, tidy=TRUE, size = "footnotesize"}
## first days of years
seq(as.Date("2007/1/1"), as.Date("2010/1/1"), "years")
## by month
seq(as.Date("2000/1/1"), by = "month", length.out = 4)
## quarters
seq(as.Date("2000/1/1"), as.Date("2001/1/1"), 
    by = "quarter")
```

## Time Intervals / Differences
- Function difftime calculates a difference of two date/time objects and returns an object of class "difftime" with an attribute indicating the units.
```{r}
time1 <- as.Date("2017-10-05")
time2 <- as.Date("2008-07-08")
time1 - time2
difftime(time1, time2, units = "weeks")
```


## Dates in Microsoft Excel
- Microsoft Excel stores dates as the number of days since Decemeber 31, 1899.
- However, Excel also incorrectly assumes that the year 1900 is a leap year to allow for compatability with Lotus 1-2-3.
- Therefore, for dates after 1901, set the origin to Decemeber 30, 1899 to convert an Excel date to an R date.
```{r}
as.Date(43013, origin = "1899-12-30")
```

# Lubridate

## Lubridate
 
- Lubridate is an `R` package that makes it easier to work with dates and times.
- Lubridate was created by Garrett Grolemund and Hadley Wickham.

```{r}
# install.packages("lubridate")
library(lubridate)
```
 

## Parse a date
 
- Lubridate accepts lots of formats

```{r}
ymd("20110604")
mdy("06-04-2011")
dmy("04/06/2011")
```
 

## Parse a date and time

```{r}
ymd_hms("2011-06-04 12:00:00", tz = "Pacific/Auckland")
```
 

## Extraction

```{r}
arrive <- ymd_hms("2011-06-04 12:00:25")
second(arrive)
second(arrive) <- 45
arrive
```
 

## Intervals

```{r}
arrive <- ymd_hms("2011-06-04 12:00:00")
leave <- ymd_hms("2011-08-10 14:00:00")
interval(arrive, leave)
```
 

## Arithmetic

```{r}
mydate <- ymd("20130130")
mydate + days(2)
mydate + months(5)
```
 

## Arithmetic
```{r}
mydate <- ymd("20130130")
mydate + days(1:5)
```
 

## End of (next) month

```{r}
jan31 <- ymd("2013-01-31")
jan31 + months(1)
ceiling_date(jan31, "month") - days(1)
floor_date(jan31, "month") + months(2) - days(1)
```


# Time Series Data in `R` (`xts`)

## What is a Time Series?
- A time series is a set of observations $x_t$, each one being recorded at a specified time $t$.

## Key `R` Time Series Packages
- `xts`: eXtensible Time Series.
- `zoo`: Z's Ordered Observations.
    - Both were created by Achim Zeileis.
- `lubridate`

<!-- ## Date Classes in `R` -->
<!-- - `Date` is in `yyyy-mm-dd` format. -->
<!-- - `POSIXct` represents the (signed) number of seconds since the beginning of 1970 (in the UTC time zone) as a numeric vector. -->
<!-- - `POSIXlt` is a named list of vectors representing `sec`, `min`, `hour`, `mday`, `mon`, `year`, time zone par maters, and a few other items. -->
<!-- ```{r} -->
<!-- x <- Sys.time()  # clock time as a POSIXct object -->
<!-- x -->
<!-- class(x) -->
<!-- ``` -->

## What is `xts`?
- `xts` is an extended `zoo` object.
- A `zoo` object is a matrix with a vector of times that form an index.
```{r, message=FALSE}
library(xts)
# xts is a matrix plus an index
x <- matrix(1:4, nrow=2, ncol=2)
idx <- seq(as.Date("2016-10-27"), length=2, by="days")
x_xts <- xts(x, order.by = idx)  
x_xts
```

## Constructing `xts`
- The function `xts()` gives you a few other options as well.
    - See `?xts`.
    - `unique` forces times to be unique.
    - `tzone` sets the time zone of the series.
- The index should be of class Date, POSIX, timeDate, chron, etc.
- If the dates are not in chronological order, the `xts` constructor will automatically order the time series.
- Since `xts` is a subclass of `zoo`, `xts` gives us all the functionality of `zoo`.

## Deconstructing `xts`
- How do we get the original index and matrix back?
    - `coredata` extracts the matrix.
    - `index` extracts the index.
```{r}
coredata(x_xts)  # Gives us a martix
index(x_xts)  # Gives us a vector of dates
```

## Viewing the **`str`**ucture of an `xts` Object.
- The `str()` function compactly displays the internal `str`ucture of an `R` object.
```{r}
str(x_xts)
```

## Importing and Exporting Time Series
- Importing:
    1. Read data into `R` using one of the usual functions.
        - `read.table()`, `read.xts()`, `read.zoo()`, etc. 
    2. `as.xts()` converts `R` objects to `xts`.
- Exporting:
    - `write.zoo(x, "file")` for text files.
    - `saveRDS(x, "file")` for future use in `R`.

## Subsetting Time Series
- `xts` supports one and two-sided intervals.
```{r}
# Load fund data
data(edhec, package = "PerformanceAnalytics")
edhec["2007-01/2007-02", 1]  # interval
head(edhec["2007-01/", 1], n = 5)  # start in January 2007
```

## Truncated Dates
- `xts` allows you to truncate dates
```{r}
# January 2007 to March
edhec["200701/03", 1]  # interval
```


## Other Ways to Extract Values
- We can subset `xts` objects with vectors of integers, logicals, or dates.
```{r}
edhec[c(1,2), 1]  # integers
edhec[(index(edhec) < "1997-02-28"), 1]  # a logical vector
edhec[c("1997-01-31","1997-02-28") , 1]  # a date vector
```

## `first()` and `last()` Functions
- `R` uses `head()` and `tail()` to look at the start and end of a series.
    - i.e. "the first 3 rows" or "the last 6 rows".
- `xts` has two functions `first()` and `last()`.
    - i.e. "the first 6 days" or "the last 6 months" 
```{r}
first(edhec[, "Convertible Arbitrage" ], "3 months")
```

## Math Operations
- Math operations are on the intersection of times.
```{r}
x <- edhec["199701/02", 1]
y <- edhec["199702/03", 1]
z <- edhec["199703/04", 1]
x + y # only the intersection
```

## Operations on the Union
```{r}
x + merge(y, index(x), fill = 0)
x + merge(y, index(x), fill = na.locf)
```


## Database Joins
- There are four main database joins: inner, outer, left and right joins.
    - inner join: intersection.
    - outer join: union.
    - left: using times from the left series.
    - right: using times from the right series.

## Merging `xts` objects
- We can merge `xts` objects using the `merge` function.
- `merge` takes three arguments.
    - an arbitrary number of time series.
    - `fill`, which handles missing data.
    - `join`, the type of join we want to do. The default is *outer* join.
```{r}
colnames(x) <- "x"; colnames(y) <- "y"; colnames(z) <- "z"
merge(x, y, z)
```

## Merging `xts` Objects: Left and Right Joins
```{r}
merge(x, y, join='left')
merge(x, y, join='right')
```

## Missing Data
- `locf`: last observation carried forward
```{r}
x <- c(1, NA, NA, 4)
idx <- seq(as.Date("2016-10-27"), length=4, by="days")
x <- xts(x, order.by = idx); colnames(x) <- "x"
cbind(x, na.locf(x), na.locf(x, fromLast = TRUE))
```
## Other `NA` Options
```{r}
na.fill(x, -999)
na.omit(x)
```

## Interpolate NAs
- Missing values (`NA`s) are replaced by *linear* interpolation via `approx` or *cubic spline* interpolation
via `spline`, respectively.


```{r}
na.approx(x)
# na.spline(x) gives the same results in this example
```

## Lagging a Time Series
- `lag(x, k = 1, na.pad = TRUE)`
    - k is the number of lags (positive = forward and negative = backward)
    - k can be a vector of lags
    - 'na.pad' pads the vector back to the original size
```{r}
x <- na.approx(x)
cbind(x, lag(x,1), lag(x,-1))
```

## Diffferencing Series
- Differencing converts levels to changes.
- see `diff.xts` for additional function arguments.
```{r}
x <- na.approx(x)
cbind(x, diff(x))
```

## Apply over Time Periods
- `period.apply()` applys a function over time intervals.
- `endpoints` gives us the row numbers of endpoints.
- `apply.monthly`, `apply.daily`, `apply.quarterly`, etc. take care of the endpoint calculation for us.
```{r}
edhec9701 <- edhec["1997/2001", c(1,3)]
# determine the endpoints
ep <- endpoints(edhec9701, "years")
period.apply(edhec9701, INDEX=ep, FUN=mean)
```

## `do.call`: A Useful `R` Trick
- The `do.call` function allows us to specify the name of function, either as a character or an object, and provide arguments as a list.
```{r}
do.call(mean, args= list(1:10))
do.call("mean", args= list(1:10))
```

## Discrete Rolling Windows
- `split`, `lapply` a function (`cumsum`, `cumprod`, `cummin`, `cummax`), and recombine.
```{r}
edhec.yrs <- split(edhec[,1], f="years")
edhec.yrs <- lapply(edhec.yrs, cumsum)
edhec.ytd <- do.call(rbind, edhec.yrs)
edhec.ytd["200209/200303", 1]
```

## Continuous Rolling Windows
- `rollapply(data, width, FUN, ...)`
```{r}
rollapply(edhec["200301/06", 1], 3, mean)
```

# Stock Market Data in `R`

## Data from `quantmod`
 
- With `quantmod` we can download stock market data into xts objects.

```{r,message=FALSE}
library(quantmod)
getSymbols("^GSPC", src="yahoo", from = "2008-01-01")
head(GSPC,3)[, 1:4]
```
 

## A Basic Plot

```{r}
plot(GSPC$GSPC.Close)
```
 

## Switch Period
 
- `to.period` changes the periodicity of a univariate or OHLC (open, high, low, close) object.

```{r}
eom <- to.period(GSPC,'months')
head(eom,3)
```

# Plotting in `R`

## Motivation
\begin{center}
\Large One skill that isn't taught in grad school is how to make a nice chart. \\
\vspace{5mm}
\normalsize - Managing Director at Citigroup
\end{center}

## What makes a chart nice?
- The reader should look at the chart and immediately understand what data are displayed.
- This means we need:
    - A clear title.
    - Clear labels for each axis (scale and units).
    - A legend if more than one time series is displayed.
    - Different colors and line formats for different time series.
    - Grid lines.
    - Labels.

## Plotting Facilities in `R`
- R has excellent plotting methods built-in.
- I will focus on base `R`.
- As a next step, I recommend learning `ggplot2`, an excellent plotting package.
- http://www.r-graph-gallery.com/

## Basic Plotting
 
- `example(plot)`
- `example(hist)`
- `?par`
- `?plot.default`

## The `plot()` Function
- `plot()` is generic function, i.e. a placeholder for a family of functions.
    - the function that is actually called depends on the class of the object on which it is called.
- `plot()` works in stages.
    - you can build up a graph in stages by issuing a series of commands.
- We will see how this works with an example.


## A Basic Plot
```{r, fig.height=5}
x <- seq(1:12)
y <- c(69, 68.7, 70.7, 73.2, 74.9, 78.2,
       82.6, 84.4, 83.5, 79, 73.6, 67.8)
plot(x, y)
```

<!-- ## `xlim()` and `ylim()` -->

## Graphical **`par`**amaters
- Graphical parameters can be set as arguments to the `par` function, or they can be passed to the `plot` function.
- Make sure to read through `?par`.
- Some useful parameters:
    - `cex`: sizing of text and symbols
    - `pch`: point type.
    - `lty`: line type.
        - 0=blank, 1=solid (default), 2=dashed, 3=dotted, 4=dotdash, 5=longdash, 6=twodash
    - `lwd`: line width.
    - `mar`: margins.
    - `bg`: background color

## `pch`
- `pch` sets how points are displayed

\centerline{
  \includegraphics[width=\textwidth,keepaspectratio]{./pch.png}
}
```{r, fig.height = 5}
plot(x,y, pch = 16, col='darkblue')
```

## Colors in `R`
- `colors()` returns all available color names.
- `rainbow(n)`, `heat.colors(n)`, `terrain.colors(n)` and `cm.colors(n)` return a vector of `n` contiguous colors.
```{r, fig.height = 5}
plot(x, y, pch = 21, col=heat.colors(12),
     cex = 2, bg = rev(heat.colors(12)))
```

## `lines()`
- `lines()` takes coordinates and joins the corresponding points with line segments.
    - Notice, by calling `lines` after `plot` the line is on top of the points.
    - This is why we want to build the plot in stages.
```{r, fig.height = 4}
plot(x,y, pch = 16, col='darkblue', cex=2)
lines(x, y, col='darkgrey', lwd = 3)
```

## `points()`
- `points` is a generic function to draw a sequence of points at the specified coordinates. The specified character(s) are plotted, centered at the coordinates.
```{r, fig.height = 4}
plot(x,y, pch = 16, col='darkblue', cex=2)
lines(x, y, col='darkgrey', lwd = 3)
points(x, y, col=rainbow(12), pch=1:12, cex=3, lwd=2)
```

## `grid()`
- `grid` adds a rectangular grid to an existing plot.
- `?grid` for more details.
```{r, fig.height = 4}
plot(x,y, pch = 16, col='darkblue', cex=2)
lines(x, y, col='darkgrey', lwd = 3)
points(x, y, col=rainbow(12), pch=1:12, cex=3, lwd=2)
grid(col="blue", lwd=2)
```

## `abline()`
- `abline` adds one or more straight lines through the current plot.
```{r, fig.height = 4}
x2 <- 1:10; y2 <- 1 + 2*x2 + rnorm(10)
plot(x2,y2, pch = 16, col='darkblue')
model <- lm(y2 ~ x2)
abline(model, col="darkgrey", lwd=2)
abline(v = 5 , col = "red", lty = 2)
abline(h = 12, col = "red", lty = 2)
```

## Adding a Title in Lables
- Use the `main` argument for a title.
- Use the `xlab` and `ylab` for axis labels.
```{r, fig.height = 5}
plot(x,y, pch = 16, col='darkblue',
     xlab = "Month", ylab = "Temperature (degrees F)",
     main = "Average High Temperature in Los Angeles, CA")
```

## Adding a Legend: The `legend()` Function
- see `?legend` and `example(legend)`
```{r, fig.height = 5}
plot(x,y, pch = 16, col='darkblue')
legend("topleft", inset=.01, "Average High Temperature",
  col = "darkblue", pch = 16, bg="white",box.col="white")
```

## `text()` and `locator()`
- Use the `text()` function to add text anywhere in the current graph.
- `locator()` allows you to click on a point in the chart and returns the location.
```{r, fig.height = 4}
plot(x,y, pch = 16, col='darkblue')
text(2,75, "ABC")
```

## `curve()`
- With `curve()`, you can plot a function.
```{r, fig.height = 5}
curve(cos(x), -2*pi, 2*pi)
```


## Saving a Plot to a File
- Saving a plot to a pdf file:
    - Open a file: `pdf("FileName.pdf", width, height,...)`
    - Create the plot.
    - Close the device with `dev.off()`
- You can use `dev.copy()` to save the displayed graph.
- See `library(help = "grDevices")` for more information.

## An Example of Plotting in `R`
- Let's plot the cumulative (gross) return of IBM and the S\&P 500 since 1980.
```{r,message=FALSE}
library(quantmod)
getSymbols(c("^GSPC", "IBM"), src="yahoo", 
           from = "1979-12-31")
adj_close <- merge(GSPC$GSPC.Adjusted, IBM$IBM.Adjusted)
daily_returns <- diff(adj_close)/lag(adj_close)
cum_ret <- cumprod(1+daily_returns[-1,])
ret1 <- xts(matrix(1, ncol=2), as.Date("1979-12-31"))
cum_ret <- (rbind(cum_ret, ret1) - 1)*100
colnames(cum_ret) <- c("GSPC", "IBM")
```


## The Data
```{r}
head(cum_ret, 9)
```

## Start with a Blank Chart and Build it Up
```{r, eval=FALSE}
plot(cum_ret$IBM, xlab="", ylab = "Cumulative Net Return (Percent)",
     main="", major.ticks="years", minor.ticks=F,
     type="n", major.format = "%Y", auto.grid=F,
     ylim = c(-500, 3000))
abline(h=seq(-500,3000,500), col="darkgrey", lty=2)
lines(cum_ret$GSPC, col="black", lwd=2)
lines(cum_ret$IBM, col="blue", lwd=2)
legend("topleft", inset=.02, 
       c("IBM","GSPC"), col=c("blue", "black"),
       lwd=c(2,2),bg="white", box.col="white")
```

## The Chart
```{r, echo=FALSE}
plot(cum_ret$IBM, xlab="",
     ylab = "Cumulative Net Return (Percent)",
     main="", major.ticks="years", minor.ticks=F,
     type="n", major.format = "%Y", auto.grid=F,
     ylim = c(-100, 3000))
title(main = "Cumulative Net Return of IBM and GSPC")
abline(h=seq(0,3000,500), col="darkgrey", lty=2)
lines(cum_ret$GSPC, col="black", lwd=2)
lines(cum_ret$IBM, col="blue", lwd=2)
legend("topleft", inset=.02, 
       c("IBM","GSPC"), col=c("blue", "black"),
       lwd=c(2,2),bg="white", box.col="white", cex = 1.25)
```

# An Example from Google Trends

## Google Trends
- Google Trends is a useful way to compare changes in popularity of certain search terms over time.
- Google Trends data can be used as a proxy for all sorts of difficult-to-measure quantities like economic activity and disease propagation. 
- Let's download data on search activity for the key word, "Data Science".
    - See the `.csv` file named `multiTimline.csv`. 

## Interest in Data Science over Time

\centerline{
  \includegraphics[width=\textwidth,height=.8\textheight,keepaspectratio]{./googletrends.png}
}

## Loading the Data into `R`

- First, we load the data into `R`:
```{r,size='scriptsize'}
dataScience = read.csv(file = "./multiTimeline.csv",
          header = T, skip = 2, stringsAsFactors = F)
colnames(dataScience) = c("date", "interest")
str(dataScience)
```

## First attempt at a Plot
- Without a x-axis, `R` treats the data as equally-spaced.
```{r, fig.height=5}
plot(dataScience$interest, type = 'l',
     ylab = "Index",xlab="",
     main = "Search Activity for `Data Science`")
```

## Convert the Date String to a Date Class
- The month column does not have a day and we need to add a day to convert it to a date.
- Let's use the first day of the month
```{r}
dataScience$date = paste0(dataScience$date, "-01")
str(dataScience)
dataScience$date = as.Date(dataScience$date,
                             format = "%Y-%m-%d")
```

## A Plot with Dates: Default Axis
- The default spacing of 5 years seems too large.
```{r, fig.height=5, size='scriptsize'}
plot(dataScience$date, dataScience$interest, type = 'l',
     ylab = "Index", xlab = "",
     main = "Search Activity for `Data Science`")
```

## A Plot with Dates: Custom Axis
```{r, fig.height=5, size='scriptsize'}
plot(dataScience$date, dataScience$interest, type = 'l',
     ylab = "Index", xlab = "",  xaxt = 'n',
     main = "Search Activity for `Data Science`")
dts = seq(from = as.Date("2004-01-01"),
           to = as.Date("2018-01-01"), by="2 years")
axis(1, at=dts, labels=format(dts, "%Y"))
```


# Lab

## Lab 2
<!-- \begin{center} -->
- \Large Let's work on [Lab 2](https://github.com/brettdunn/mfeR2017/blob/master/week4/lab2/Lab2.pdf).
<!-- \end{center} -->