---
title: "MFE R Programming Workshop"
author: "Mahyar Kargar and Dan Yavorsky"
date: "Fall 2018"
output:
  beamer_presentation:
    includes:
      in_header: ../slide_header.tex
    keep_tex: false
subtitle: Week 3
slide_level: 2
---
```{r, include=FALSE}
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
```


# Introduction

## Questions
\begin{center}
\Large Any questions before we start?
\end{center}

## Overview of Week 3
- We will cover four classes to store array-like data:
    - Matrices
    - Data Frames
    - Tibbles
    - Intro to Data Tables

## Changing the Working Directory
- Use the `setwd()` R function
- Use the **Tools | Global Options | General**  menu 
- From within the Files pane, use the **More | Set As Working Directory** menu. (Navigation within the Files pane alone will not change the working directory.)

## Matricies, `data.frame`s, `tibble`s, and `data.table`s
- Matrix is a data type in R with the dimension attribute - the rows and the columns.
    - It has the elements of *same* class type.
    - We can have character, integer or complex elements in the matrices and so on.
    - We *cannot* have elements of mixed modes/class types such as both integer and character elements in the same matrix.
- A `data.frame` is a list of vectors of equal length, and the vectors can be of different types.
    - e.g. one character column, one numeric column.
- Tibbles and data tables inherit the functionality of data frames and improve on them in various ways.

# Matrices

## Creating Matrices
- Matrices are vectors with a number of rows and number of columns attribute.
```{r}
myvec <- 1:10
mymat <- matrix(myvec, nrow=2, ncol=5, byrow = FALSE)
mymat
dim(mymat)  # returns the dimension
```

## Accessing Elements of Matrices
- Like vectors, elements can be accessed using `[]`
```{r}
mymat <- matrix(1:15, nrow=3, ncol=5)
mymat[1, 2]  # row 1, column 2
mymat[2:3, c(1, 4, 5)]
```

## Filtering Matrices
- Filtering can be done on a single column or a single row, otherwise the filter returns a vector.
```{r}
myvec <- c(1, 1, 3, 1, 5, 1, 7, 1, 9, 1)
mymat <- matrix(myvec, nrow=2, ncol=5)
mymat
mymat[, mymat[1, ] > 4]
```

## Vectorization
- Most R functions work on matrices as well.
```{r}
mymat <- matrix(1:10, nrow=2, ncol=5)
exp(mymat)
sd(mymat)  # standard deviation
```

## Applying Functions to Rows and Columns
- `apply` allows you to apply a function across a dimension of a matrix.
- The third argument is a function!
```{r}
mymat <- matrix(1:10, nrow=2)
# mean across rows (can also use rowMeans() function)
apply(mymat, 1, mean)  # apply mean along rows
apply(mymat, 2, max)  # apply max along columns
```


## Combining Matricies with `cbind` and `rbind`
- Column bind and row bind.
```{r}
mymat1 <- matrix(1:4, nrow=2)
mymat2 <- matrix(6:9, nrow=2)
mymat3 <- matrix(10:11, ncol=2)
cbind(mymat1, mymat2)
rbind(mymat1, mymat3)
```

## Matrix operations
- Many matrix operations are surrounded by `%` signs.
```{r}
mymat1 <- matrix(1:4, nrow=2)
mymat2 <- matrix(5:8, nrow=2)
mymat1 %*% mymat2  # matrix multiplication
mymat1 + mymat2
```

## Matrix Algebra
- http://www.statmethods.net/advstats/matrix.html

# Data Frames

## `data.frame`s
- The `data.frame` is one of the most useful features in R.
- A `data.frame` is like a `matrix` with a two-dimensional rows-and-columns structure.
- However, unlike a `matrix`, in a `data.frame` each column can have a *different* data type.
    - For example, one column might be numbers and another characters.
- Technically, a `data.frame` is a `list`, with the components of the list being equal-length vectors.
- Each column must be the same length (unlike a list).

## Creating `data.frame`s
- Unless you are working with categorical data, you probably want to set `stringsAsFactors=FALSE`.
```{r}
courses <- c("Stochastic Calculus", "Fixed Income")
examGrades <- c(92, 98)
gradeBook <- data.frame(courses, examGrades, 
                        stringsAsFactors = FALSE)
gradeBook
```

## Column Names
- Column names in data.frames are specified by `names()`.
- This is because data.frames are actually lists with special attributes.
- That means that the usual list functions work on data.frames.
- `lapply`, etc.

# Accessing Elements of `data.frames`

## Accessing Elements of `data.frame`s
- We can access a `data.frame` component just like a `list`.
```{r}
gradeBook[[1]]           # first way
gradeBook[["courses"]]   # second way
gradeBook$courses        # third way
```

## Accessing Elements of `data.frame`s (2)
- Note that `[[` simplifies the result, returning a vector:
```{r}
str(gradeBook[[1]])
```
- But `[` is preserving subsetting and (usually) returns a `data.frame:
```{r}
str(gradeBook[1])
```
## Accessing Elements of `data.frame`s like a `matrix` 
- We can access `data.frame` elements like a `matrix`.
```{r}
gradeBook[1,2]
gradeBook[1,] # the first row
gradeBook[,2] # the second column
```

## Accessing Elements of `data.frame`s like a `matrix`(2)
- To preserve the `data.frame` class, set `drop = FALSE`
```{r}
gradeBook[,2] # returns a vector
gradeBook[,2,drop=FALSE] # returns a data.frame
```

# Subsetting `data.frames`
## Filtering with `subset()`
```{r}
set.seed(1234)
x.df <- data.frame(V1 = rnorm(4),V2 = runif(4), 
                   V3 = rchisq(4,df = 2), V4 = 1:4)
x.df

x.sub <- subset(x.df, V4 > 2)
x.sub
```

## Subsetting rows using conditional statements
- The data frame `x.sub1` contains only the observations for which the values of the variable `V4` is greater than 2 and the variable `V1` is greater than 0.6.
```{r}
x.sub1 <- subset(x.df, V4 > 2 & V1 > 0.6)
x.sub1
```

## Subsetting both rows and columns
- The data frame `x.sub2` contains only the variables `V2` and `V3` and then only the observations of these two variables where the values of variable `V4` are greater than 2 and the values of variable `V1` are greater than 0.6.
```{r}
x.sub2 <- subset(x.df, V4 > 2 & V1 > 0.6, 
                 select = c(V2, V3))
x.sub2
```

## Subsetting rows using indices
- The `x.sub3` data frame contains only the observations for which the values of variable `V4` are equal to 2. 
```{r}
x.sub3 <- x.df[x.df$V4 == 2, ]
x.sub3 
```

## Subsetting rows using `%in%`
- The `x.sub4` data frame contains only the observations for which the values of variable `V4` are equal to either 1 or 4.
```{r}
x.sub4 <- x.df[x.df$V4 %in% c(1, 4), ]
x.sub4
```

## Subsetting columns using indices
- The `x.sub5` data frame contains all the rows on `x.df`, removing the first and third columns
```{r}
x.sub5 <- x.df[, -c(1,3)]
x.sub5
```


<!-- ## Filtering with logicals -->
<!-- - Subsetting `data.frame` often subsetting the rows and/or columns, which are each vectors. -->
<!-- - We can select subsets of vectors with vectors of logicals. -->
<!-- ```{r} -->
<!-- x <- 1:5 -->
<!-- y <- c(TRUE,FALSE,TRUE,FALSE,TRUE) -->
<!-- x[y] -->
<!-- ``` -->

<!-- ## Filtering with logicals (2) -->
<!-- - Filtering amounts to generating filtering indices (i.e. vectors of logicals). -->
<!-- ```{r} -->
<!-- x <- c(5, 2, -3, 8) -->
<!-- idx <- x*x > 8  # same as ">"(x*x, 8); 8 is recycled! -->
<!-- x[idx] -->
<!-- ``` -->

<!-- ## Assigning to a Filter -->
<!-- - You can assign elements to the subsets. -->
<!--     - This allows you change elements that meet certain criteria. -->
<!-- ```{r} -->
<!-- x <- 1:6 -->
<!-- x[x < 2] <- NA -->
<!-- x -->
<!-- ``` -->

<!-- ## Filtering with `subset()` -->
<!-- - The subset function filters and removes any `NA`s. -->
<!-- ```{r} -->
<!-- x <- c(3, 1:5, NA, 79) -->
<!-- x -->
<!-- x[x > 4] -->
<!-- subset(x, x > 4) -->
<!-- ``` -->
<!-- ## Filtering with Indices -->
<!-- - Select using the row/column number. -->
<!-- - Exclude entries with a negative vector. -->
<!-- ```{r} -->
<!-- x <- 10:20 -->
<!-- x[c(1:2, 6:7)] -->
<!-- x[-c(1,3)] # removes 1st and 3rd obs -->
<!-- ``` -->
<!-- ## The Selection Function `which()` -->
<!-- - `which()` gives us the position in a vector where a condition occurs. -->
<!-- ```{r} -->
<!-- x <- c(3, 1:5, NA, 79) -->
<!-- x > 4 -->
<!-- which(x > 4) -->
<!-- ``` -->
<!-- - See https://stackoverflow.com/questions/6918657/whats-the-use-of-which. -->

<!-- ## `%in%` -->
<!-- - `%in%` returns a logical vector indicating if there is a match or not for its left operand. -->
<!-- ```{r} -->
<!-- 1:5 %in% c(1,3) -->
<!-- ``` -->

<!-- ## Filtering Matrices -->
<!-- - Filtering can be done on a single column or a single row, otherwise the filter returns a vector. -->
<!-- ```{r} -->
<!-- myvec <- c(1, 1, 3, 1, 5, 1, 7, 1, 9, 1) -->
<!-- mymat <- matrix(myvec, nrow=2, ncol=5, byrow = FALSE) -->
<!-- mymat -->
<!-- mymat[, mymat[1, ] > 4] -->
<!-- ``` -->

## Complete.cases
- `complete.cases()` gets rid of any *rows* with at least one `NA` value.
```{r}
# Let's makes the second col in row 1 an NA
x.df[1,2] <- NA              
x.df[complete.cases(x.df), ] # removes row 1
```

## Merging `data.frame`s
- Two `data.frame`s can be combined using the `merge` function.
```{r}
courses <- c("Stochastic Calculus", "Fixed Income")
midtermGrades <- c(89, 91)
gradeBook2 <- data.frame(courses, midtermGrades,
                         stringsAsFactors = FALSE)
merge(gradeBook, gradeBook2)
```

## Adding Columns to `data.frame`s
```{r}
dat1 <- 1:4
dat2 <- rep(c("A","B"),each=2)
myframe <- data.frame(col1=dat1,col2=dat2)
myframe$col3 <- 5:8
myframe
```

## Reading in Data from a CSV File
- Reading in data typically gives you a `data.frame`.
- `read.table` is the basic function to read in tabular data.
- `read.csv` is a special case of `read.table`.
- As usual see `?read.table`.
- Often you want to set `stringsAsFactors = FALSE`.
- `write.csv` writes data to a .csv file.
```{r}
optdata <- read.csv(file="../week2/lab/optionsdata.csv",
                header = T, stringsAsFactors = FALSE)
head(optdata, 3)
```

# `tibble`s

## What is a `tibble`?
- Tibbles are a "modern take" on R's traditional `data.frame`.
- They keep the features that have stood the test of time, and drop the features that used to be convenient but are now frustrating (i.e. converting character vectors to factors).
```{r,echo=FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```


```{r,tidy = TRUE, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
as.tibble(iris)
```

## Creating a `tibble`
- `tibble()` can be used to create a data frame.
- It never changes an input’s type (i.e., no more `stringsAsFactors = FALSE`!).
```{r}
tibble(x = letters)
```
## Creating a tibble of lists
This makes it easier to use with list-columns:
```{r}
tibble(x = 1:3, y = list(1:5, 1:10, 1:20))
```

## Lazy and Sequential Evaluation
It evaluates its arguments lazily and sequentially:
```{r}
tibble(x = 1:5, y = 1, z = x ^ 2 + y)
```

## Column Names
- Tibbles never adjust the names of variables:
```{r}
names(data.frame(`crazy name` = 1))
names(tibble(`crazy name` = 1))
```

## Other Features
- Tibbles don't use `row.names()`. It never stores a variable as special attribute.
- Tibbles only recycle vectors of length 1. This is because recycling vectors of greater lengths is a frequent source of bugs.
- Tibble provides `as_tibble()` to coerce objects into tibbles.

## Tibbles vs Data Frames: Printing
- When you print a tibble, it only shows the first ten rows and all the columns that fit on one screen. It also prints an abbreviated description of the column type.
- You can control the default appearance with options.
```{r}
options(tibble.print_max = 3, tibble.print_min = 2)
tibble(x = 1:1000)
```

## Tibbles vs Data Frames: Subsetting
- If you want to pull out a single variable, you could use `$` and `[[` or `%>%` (`[[` can extract by name or position; `$` only extracts by name)
```{r}
df <- tibble( x = 1:5, y = rnorm(5))
# Extract by name (df[["x"]] works the same way)
df$x

# Extract by position
df[[1]]

df %>% .$x # df %>% .[["x"]] works the same way

```

<!-- - Recall that for data frames: -->
<!--     - `[[` is simplifying subsetting and returns a vector.  -->
<!--     -`[` is preserving subsetting and (often) returns a `data.frame`. -->
<!-- - For `tibble`s `[` always returns another `tibble`. -->

```{r, echo=FALSE, eval=FALSE}
df <- data.frame(x = 1:3, y = 3:1)
tbl <- tibble(x = 1:3, y = 3:1)
class(df[, 1])
class(tbl[, 1])
```

## Tibbles vs Data Frames: Subsetting with `[[` and `$`
- Recall that for data frames:
    - `[[` extracts a single column as a vector.
    - `$` works similarly to `[[`, but does *partial matching* on the column name.
```{r}
df <- data.frame(colName = 1:5, m = 2:6); df$c
```
- Tibbles never do partial matching, and will throw a warning and return `NULL` if the column does not exist.
```{r}
tbl <- as_tibble(df); tbl$c
```

## Tibbles vs Data Frames: Recycling
- When constructing a `tibble`, only values of length 1 are recycled.
- The first column with length different to one determines the number of rows in the tibble, conflicts lead to an error.
    - `tibble(a = 1:3, c = 1:2)` gives "`Error: Column c must be length 1 or 3, not 2`".
<!-- - This also extends to tibbles with zero rows, which is sometimes important for programming. -->
<!-- ```{r} -->
<!-- tibble(a = integer(), b = 1) -->
<!-- ``` -->

# Very quick intro to `data.table`s

## What is a `data.table`?
- Think of `data.table` as an advanced version of `data.frame`.
    - Every column is the same length, but may have a different type
- It inherits from data.frame and works even when `data.frame` syntax is applied on data.table
- `data.table` is *very fast*.
- It is one of the most useful packages in R.
- The syntax of `data.table` is concise.
    - Lowers programmer time...
    - ...but it can be hard to understand
    - Make sure you comment your code!
```{r, message = FALSE}
library(data.table)
```

## An Example
- Syntax is `DT[i, j, by]`: 
  - "Take DT, subset rows using `i`, then calculate `j` grouped by `by`."

```{r, message = FALSE}
data("mtcars")
mtcarsDT <- data.table(mtcars)
mtcarsDT[ mpg > 20, 
          .(AvgHP = mean(hp), "MinWT(kg)" = min(wt*453.6)), 
          by = .(cyl, under5gears = gear < 5)]

```

## Why learn `data.table`?
- For data that fits in memory, `data.table` is much faster than `data.frame`s and `tibble`s.
- `data.table` also saves memory because it avoids copying large objects.
- Part of the speed advantage comes from the fact that `data.table` provides a set of tools to update by reference.
    - In base `R`, if a function modifies a single element of a large `data.frame`, a copy of the entire `data.frame` is made.
- `data.table` provides a powerful set of commands to access data.
- We will cover `data.table` in much more detail later in the course.