---
title: "MFE R Programming Workshop"
author: "Brett Dunn and Mahyar Kargar"
date: "Fall 2017"
output:
  beamer_presentation:
    includes:
      in_header: ../slide_header.tex
    keep_tex: false
subtitle: Week 5
slide_level: 2
---
# Introduction

## Questions
\begin{center}
\Large Any questions before we start?
\end{center}

## Overview
- Importing Data from Databases
- Importing Data from the Web
- Importing Data from WRDS
- `data.table`

<!-- # Random R Tricks -->

<!-- ## Send an Email from R -->
<!-- ```{r, eval = FALSE} -->
<!-- library(gmailr) -->
<!-- hiFromR <- mime( -->
<!--   To = "someone@anderson.ucla.edu", -->
<!--   From = "me@gmail.com", -->
<!--   Subject = "hello from R", -->
<!--   Body = "I am writting you from R." -->
<!-- ) %>% -->
<!-- attach_file("./file.R") -->
<!-- send_message(hiFromR) -->

<!-- ``` -->

<!-- ## Working with Files in R -->
<!-- ```{r, eval = FALSE} -->
<!-- dir.exists("week3") # does the directory exist? -->
<!-- dir.create("week4") # create a directory -->
<!-- # download files -->
<!-- download.file(url, destinationfile, method='curl') -->
<!-- untar()  # untar a file -->
<!-- tar()  # tar a file -->
<!-- # display file information -->
<!-- f <- (dir("/path/to/something",  -->
<!--         full.names = TRUE) %>% file.info) -->
<!-- # delete a file (be careful!!) -->
<!-- unlink("/path/to/something") -->

<!-- ``` -->

# Importing Data from Databases

## Databases
- A database stores data in table format.
- There a several popular database management systems (DMS):
    - MySQL, PostgreSQL, SQLite, Oracle, Microsoft SQL Server, etc.
- Structured Query Language (SQL) is used maintain and query the database.

## Databases in R
- There are different R packages for each DMS:
    - MySQL = RMySQL
    - PostgreSQL = RPostgreSQL
    - SQLite = RSQLite
    - Oracle = ROracle
    - SQL Server = RSQLServer
- The functions we use to interact with the database are specified in the R package `DBI`.

## A MySQL Example: Establish a Connection
- First, we need to establish a connection to the database.
```{r, eval=FALSE}
library(DBI)
con <- dbConnect(RMySQL::MySQL(), # the MySQL driver
                 dbname = "dbname", # database name
                 host = "IP_or_WebAddress",
                 port = 3306, # port is 3306 by default
                 user = "username",
                 password = "password")
```

## List and Import Tables
```{r, eval=FALSE}
# lists the tables in the database
dbListTables(con)
# returns a dataframe
dbReadTable(con, "tablename")
# import all the tables
table_names <- dbListTables(con)
tables <- lapply(table_names, dbReadTable, conn = con)
# close the connection when you are done
dbDisconnect(con)
```

# Importing Data From the Web

## JSON
- JSON object: an unordered collection of name-value pairs.
- JSON array: an ordered sequence of zero or more values.
- JSON objects and arrays can be nested in each other.
- R handles JSON with the `jsonlite` package.

## An Example with `jsonlite`
- Let's get the current wind and temperature status at LAX.
```{r}
library(jsonlite)
airportCode <- "LAX"
url <- paste0("http://services.faa.gov/airport/status/",
              airportCode)
LAX <- fromJSON(url)
LAX$weather$wind
LAX$weather$temp
```

## Quandl
- Quandl is a useful source of financial data and there is an R package `Quandl` to import the data into R.
- See https://www.quandl.com/tools/r.
- Data can be downloaded as `xts` objects, datatables, etc.
```{r, message=FALSE}
library(Quandl)
# download GDP % growth as an xts object
gdp <- Quandl("FRED/GDP", type="xts")
last(gdp, 4)
```

# Importing Data from WRDS

## WRDS, CRSP, and R
- Wharton Research Data Services (wrds) has over 250 terabytes of data.
- One data provider is The Center for Research in Security Prices (CRSP).
    - You will use CRSP data throughout the MFE program.
- I will show you how to access WRDS from R.
- Documentation: [Using R with WRDS](https://wrds-web.wharton.upenn.edu/wrds/support/Accessing%20and%20Manipulating%20the%20Data/_007R%20Programming/_001Using%20R%20with%20WRDS.cfm)

## Setup
- First, you need to encode your wrds password: instructions [here](https://wrds-web.wharton.upenn.edu/wrds/support/Accessing%20and%20Manipulating%20the%20Data/_010Encoding%20your%20WRDS%20Password.cfm).
- We also need to obtain access to WRDS and download the SAS drivers for JDBC from [here](https://wrds-web.wharton.upenn.edu/wrds/support/SAS-JDBC-Drivers.zip.cfm).
- The two files should be saved locally.
- Take note of the path to the files; we need the path to establish the connection to WRDS.

## Establish the Connection
```{r eval=FALSE, include=FALSE}
library(rJava)
.jinit(parameters="-Xmx8g")
library(RJDBC)

user <- "my_username"
pass <- '{SAS002}DBCC5712369DE1C65B19864C1564FB850F398DCF'
wrdsconnect <- function(user=user, pass=pass){
drv <- JDBC("com.sas.net.sharenet.ShareNetDriver", "~/Dropbox/wrds_driver/sas.intrnet.javatools.jar",
    identifier.quote="`")
wrds <- dbConnect(drv, "jdbc:sharenet://wrds-cloud-h.wharton.private:8551/", user, pass)
return(wrds)
}
.First <- function(){
    wrds <<- wrdsconnect(user=user, pass=pass)
}
.Last <- function(){
    dbDisconnect(wrds)
}
```

```{r eval=FALSE}
# ---- INPUTS ---- #
username <- "myUserName"
password <- "myPassword"
# local path to the sas files
sasPath <- "C:/Users/myUser/Documents/wrds-drivers"
# ---- CODE ---- #
library(rJava)
options(java.parameters = '-Xmx4g')
library(RJDBC)
sasCore <- paste0(sasPath, "/sas.core.jar")
sasDriver <- paste0(sasPath, "/sas.intrnet.javatools.jar")
.jaddClassPath(c(sasCore, sasDriver))
driver <- RJDBC::JDBC(
         "com.sas.net.sharenet.ShareNetDriver",
          sasDriver, identifier.quote = "`")
wrds <- RJDBC::dbConnect(driver,
"jdbc:sharenet://wrds-cloud.wharton.upenn.edu:8551/",
 username, password)
```

## Accessing Data
- On the previous slide, we created the connection `wrds`.

```{r, eval = FALSE}
res <- dbSendQuery(wrds, "select * from DATASET")
data <- fetch(res, n = -1)
data
```
- **`dbSendQuery()`** uses `wrds` to submit the SQL query string to WRDS, which then returns the result `res`.
- **`select * from DATASET`** is a SAS SQL query.
    - See the [SAS SQL Documentation](http://support.sas.com/documentation/cdl/en/sqlproc/63043/PDF/default/sqlproc.pdf) for more information.
- **`fetch()`** fetches the actual data based on the result `res`.
- **`n = -1`** is a parameter that determines how many observations to download.
    - ``n = -1`` specifies that we'd like unlimited observations returned.
    - `n = 10` would limit the number of observations returned to 10.

## Example: S&P 500 Returns
```{r, eval = FALSE}
sql <- "SELECT caldt, vwretd FROM CRSPQ.MSP500"
res <- dbSendQuery(wrds, sql)
dbHasCompleted(res) #check that this is true
msp500 <- fetch(res, n = -1)  
dbClearResult(res) # free up memory
msp500$caldt <- as.Date(msp500$caldt)
library(xts)
msp500 <- xts::xts(msp500[, -1],
                   order.by = msp500$caldt)
colnames(msp500) <- "vwretd"
```

# `data.table`

## What is a `data.table`?
- Think of `data.table` as an advanced version of `data.frame`.
    - Every column is the same length, but may have a different type
- It inherits from data.frame and works perfectly even when data.frame syntax is applied on data.table.
- `data.table` is very fast.
- The syntax of `data.table` is very concise.
    - Lowers programmer time...
    - ...but it can be hard to understand
    - Make sure you comment your code!
- Highly recommend going through [`data.table` Cheat Sheet](https://s3.amazonaws.com/assets.datacamp.com/img/blog/data+table+cheat+sheet.pdf).
```{r, message = FALSE}
library(data.table)
```

## An Example
- Syntax is `DT[i, j, by]`
- Take DT, subset rows using **i**, then calculate **j** grouped by **by**.
```{r, message = FALSE}
data("mtcars")
mtcarsDT <- data.table(mtcars)
mtcarsDT[
  mpg > 20,
  .(AvgHP = mean(hp),
    "MinWT(kg)" = min(wt*453.6)),
  by = .(cyl, under5gears = gear < 5)]
```

<!-- ## Types in R -->
<!-- - Character columns are very efficient in R. -->
<!-- ```{r} -->
<!-- typeof(1) -->
<!-- typeof(1L) -->
<!-- typeof(NA) -->
<!-- typeof(NA_integer_) -->
<!-- ``` -->

## Creating a `data.table`
```{r, warning=FALSE}
set.seed(1234)
DT <- data.table(A=1:6, B=c("a", "b", "c"),
                 C=runif(6), D=FALSE)
DT
```

## Selecting Rows by Number in **i**
- The comma is optional.
```{r, warning=FALSE}
DT[2:4, ]
DT[2:4]
```

## Selecting Rows: other methods
- The comma is optional.
```{r, warning=FALSE}
DT[ A %in% 2:3,]
DT[ B == "a", ]
```

## Selecting Columns in **j**
- Columns are specified as a list with the actual names, not as character vectors.
- `.()` is an alias to `list()` in data.tables.
```{r, warning=FALSE}
DT[2:3, list(A, C)]
DT[2:3, .(A, C)]
```

## Selecting Columns in **j** with character vectors
- To select columns with a character vector, use the `with = FALSE`. 
```{r, warning=FALSE}
DT[2:3, c("A", "C"), with = FALSE]
```

## Computing on Columns
- If the lengths of the results are not equal, the shorter one will be recycled.
```{r}
DT[A %in% 2:4, .(Total = sum(A), Mean = mean(C))]
DT[A %in% 2:4, .(B, Mean = mean(C))]
```

## You can put almost anything into **j**
```{r, fig.height=4}
DT[, plot(A, C)]
```

## Multiple Expressions Go in Curly Braces
```{r, fig.height=4}
DT[, {print(A)
      hist(C)
      NULL}] # set return value to NULL
```

## Returning a Vector.
```{r}
DT[, .(A)]  # a data.table
DT[, A]  # a vector
```

## Doing **j** by Group

```{r}
DT[, .(Total = sum(A),
     Mean = mean(C)),
     by = "B"] # returned in the order they appear
# functions work as well
DT[, .(Total = sum(C)), by = .(Group = A%%2)]
```

## `.N`
- `.N`, when used inside square brackets, contains the number of rows.
- When put in **j**, `.N` counts the observations in each group.
```{r}
DT[.N] # the last row, the same as DT[nrow(DT)]
DT[, .(Total = sum(C), Count = .N), by = .(Group = A%%2)]
```

<!-- ## Operations can be Chained Together -->
<!-- ```{r} -->
<!-- DT[,`:=`(E=A^2, F = C - 1)][,G:=sqrt(abs(F))] -->
<!-- ``` -->


<!-- ## Operations can be Chained Together -->
<!-- ```{r, fig.height=3} -->
<!-- DT[, {plot(A, C) -->
<!--       DT}][, .(Total = sum(C), # DT returns DT -->
<!--                Count = .N), -->
<!--                by = .(Group = A%%2)] -->
<!-- ``` -->

## **`S`**ubset of **`D`**ata - `.SD`
- `.SD` is a data.table.
- `.SD` holds all the columns except for the one specified in **by**, and `.SD` is only accessible in **j**.
- This is very useful if you have a very wide data.table.
- `.SDcols` allows you to apply a function to a subset of the columns.
```{r}
DT <- as.data.table(mtcars)
DT[, lapply(.SD, median), by = cyl, 
   .SDcols = c("mpg","gear","wt")]
```
- Since `lapply` returns a list, we don't need to wrap it in `.()`.

## Add or Update Columns by Reference Using `:=`
```{r}
(DT <- data.table(A=1:3, B=4:6))
DT[,C := A + B]
DT
```

## Create Multiple Columns with `:=` in One Statement
```{r, tidy=TRUE, size="footnotesize"}
DT_cars <- data.table(mtcars)[,.(mpg, cyl)]
DT_cars[,`:=`(avg=mean(mpg), med=median(mpg)), 
   by=cyl]
head(DT_cars)
```


## Remove Columns Using `:=`
- We use `NULL` to remove columns.

```{r}
DT[, D := 10:12]
DT
DT[,`:=`(B = NULL, C = NULL)]
# DT[, c("B", "C") := NULL] # also works
DT
```
- Wrap the LHS in `()` if you pass a variable.

## Combining `:=` with **i** and **by**
```{r}
DT <- data.table(A=1:6, B=c("a", "b", "c"), C=runif(6))
DT[, D := sum(C), by = B]
DT
```

## Use `set()` in Loops.
- `set()` is a loopable, low-overhead version, of the `:=` operator, but it cannot handle grouping.
- Syntax: `set(DT, i, j, value)`.
- Instead of `for (i in 1:6) DT[i, C := i+1]` we can 
```{r}
DT <- data.table(A = 7:12, B = 10:15)
for (i in 1:6) set(DT, i, "B", i+1)
# for (i in 1:6) set(DT, i, 2L, i+1) # would work too
DT
```

## `setnames()` to Change the Column Names
- `setnames(DT, "old", "new")` changes the column names by reference (no copies are being made).
```{r}
setnames(DT,c("A", "B"),c("X", "Y"))
DT
```

## `setcolorder()` Reorders the Columns by Reference
```{r}
setcolorder(DT,c("Y", "X"))
DT
```

<!-- ## Using Column Names in **i** -->
<!-- ```{r} -->
<!-- DT <- data.table(A=1:6, B=c("a", "b", "c"), -->
<!--                  C=runif(6), D=FALSE) -->
<!-- DT[B %in% c("a","b")] -->
<!-- ``` -->

## Regular Expressions
- Metacharacters allow you to match certain types of characters.
    - For example, "`.`" means any single character, "`^`" means "begins with", and "`$`" means "ends with".
- If you want to use any of the metacharacters as actual text, you need to use the \\ escape sequence.
- See `?gsub()` and `?grep()`.
```{r}
iris_dt <- as.data.table(iris)
# Change column names
setnames(iris_dt, names(iris_dt), 
         gsub("^Sepal\\.", "", names(iris_dt)))
# Remove columns
iris_dt[, grep("^Petal", names(iris_dt)) := NULL]
head(iris_dt, n = 2)
```

## Keys
- Setting a key sorts the table by the column specified.
```{r}
DT <- data.table(A=c("c", "b", "a"),B=1:6)
setkey(DT, A)
DT
```

## Keys as Row Names
- Keys can be used like row names.
```{r}
DT["a"]
DT["a", mult = "first"]
DT["a", mult = "last"]
```

## `nomatch`
- Keys can be used like row names.
```{r}
DT[c("a","d")]
DT[c("a","d"), nomatch = 0]
```

## Multi-Column Keys
- Use `.()` to select rows.
```{r}
DT <- data.table(A=c("c", "b", "a"),B=1:6,C=7:12)
setkey(DT, A, B)
DT[.("b")]
DT[.("b", 5)]
```

## `merg`ing `data.table`s
- Fast merge of two `data.table`s. It behaves very similarly to that of `data.frame`s except that, by default, it attempts to merge
    - at first based on the shared **key** columns, and if there are none,
    - then based on key columns of the first argument `x`, and if there are none,
    - then based on the common columns between the two `data.table`s.
- Set the `by`, or `by.x` and `by.y` arguments explicitly to override this default.
- Set the `all.x` (for left joins), `all.y` (for right joins), and `all` (for outer joins) logical arguments to override the default (inner joins).

## `merge` example
```{r}
(x <- data.table( foo = 1:4, a=20:23, zoo = 5:2 ))
(y <- data.table( foo = 2:4, b=30:32, boo = 10:12))
setkey(x, foo)
setkey(y, foo)
```

## `merge` example
```{r}
merge(x,y)
merge(x,y, all.x = TRUE)
```



## Using `shift` for to lead/lag vectors and lists
```{r}
DT <- data.table(mtcars)[,.(mpg)]
DT[,mpg_lag1:=shift(mpg, n = 1)]
DT[,mpg_forward1:=shift(mpg, n = 1, type='lead')]
head(DT)
```

## Reshaping `data.table`s
- The `melt` and `dcast` functions for `data.table`s are extensions of the corresponding functions from the [`reshape2`](https://cran.r-project.org/web/packages/reshape2/index.html) package.
- See the `data.table` reshape [vignette](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-reshape.html).

```{r}
DT <- readRDS("melt_example.RDS")
DT
```

## `melt`ing `data.table`s (wide to long)
```{r}
(DT.m1 <- melt(DT, 
  id.vars = c("fam_id", "age_mom"),
  measure.vars = c("dob_child1", "dob_child2", "dob_child3")))
```

## `melt`ing `data.table`s (wide to long)
- Can name the `variable` and `value` columns
```{r}
(DT.m2 <- melt(DT, 
  measure.vars = c("dob_child1", "dob_child2", "dob_child3"),
  variable.name = "child", value.name = "dob"))
```


## Casting `data.table`s (long to wide)
-  We can get back to the original `data.table` `DT` from `DT.m1` or `DT.m2`
    - collect all `child` observations corresponding to each `fam_id`, `age_mom` together under the same row. 
```{r}
dcast(DT.m2, fam_id + age_mom ~ child, value.var = "dob")

# using DT.m1
# dcast(DT.m1, fam_id + age_mom ~ variable, 
#       value.var = "value")
```

## Pass a function to aggregate by in `dcast`
- Can do that with the argument `fun.aggregate`. 
    - get the number of children in each family
```{r}
dcast(DT.m2, fam_id ~ ., 
      fun.agg = function(x) sum(!is.na(x)),
      value.var = "dob")
```

- Check `?dcast` for other useful arguments.

## `fread` and `fwrite`
- `fread` is similar to `read.csv()` but a lot faster! It reads a csv file into a `data.table`.
- `fwrite` is to write a data.table into a csv file similar to `write.csv()`.

## Converting `xts` objects to `data.table`s
```{r, message=FALSE}
library(xts)
x <- matrix(1:4, nrow=2, ncol=2)
idx <- seq(as.Date("2016-10-31"), length=2, by="months")
x_xts <- xts(x, order.by = idx)
x_xts
colnames(x_xts) <- c("a", "b")
DT <- as.data.table(x_xts)
setkey(DT,index)
DT
```

<!-- ## Rolling Joins -->
<!-- - Rolling joins are useful for time-series data. -->
<!-- - See `roll` and `rollends` in `?data.table`. -->
<!-- ```{r, message=FALSE} -->
<!-- DT -->
<!-- dt <- as.Date("2016-11-15"); DT[.(dt)] -->
<!-- DT[.(dt), roll=TRUE] # roll forward; try roll=-Inf.  -->
<!-- ``` -->

## Rolling Joins
- Suppose you have a table of product sales and a table of commercials. 
- You might want to associate each product sale with the most recent commercial that aired prior to the sale. 
<!-- - You cannot do a basic `join` between the sales and the commercials `data.table`s because each sale was NOT tracked with a `commID` attribute -->
```{r}
sales <- data.table(saleID=c("S1","S2","S3","S4","S5"), 
          saleDate=as.Date(c("2014-2-20","2014-5-1",
                "2014-6-15","2014-7-1","2014-12-31")))
sales
```

## Rolling Joins
```{r}
commercials <- data.table(commID=c("C1","C2","C3","C4"), 
               commDate=as.Date(c("2014-1-1","2014-4-1",
                                 "2014-7-1","2014-9-15")))
commercials
setkey(sales,saleDate)
setkey(commercials,commDate)

```
## Rolling Joins
```{r}
sales[commercials, roll=TRUE]
```

- `data.table` is associating each commercial with the most recent sale prior to the commercial date (and including the commercial date). 
- The most recent sale prior to each commercial is said to roll forwards, and the saleDate is mapped to the `commDate`. 
- Notice that sale `S4` was the most recent sale prior to commercials `C3` and `C4`, so `S4` appears twice in the resultant table.
- Can do *backward* rolling joins by setting the `roll = -Inf`.

## Lab 3
\begin{center}
\Large Let's work on \href{https://github.com/brettdunn/mfeR2017/blob/master/week5/Lab/Lab3.pdf}{Lab 3}.
\end{center}


